---
title: "Sample IDs and Relative Abundance Misc."
output:
  workflowr::wflow_html:
    toc: true
editor_options:
  chunk_output_type: console
---

*	Sampling IDs in each dataset with Fusobacterium (genus level) abundance above 0.1%
* I need to know the relative abundance of Streptococcus sanguinis, Prevotella (genus level), Fusobacterium nucleatum, Camplybactor concisus in the NCI-MD dataset for tumors only.


```{r data, include=FALSE, echo=FALSE}
# load packages
source("code/load_packages.R")

# get data
source("code/get_cleaned_data.R")

theme_set(theme_bw())
pal = "Sequential"
scale_colour_discrete <-  function(palname=pal, ...){
  scale_colour_brewer(palette=palname, ...)
}
scale_fill_discrete <-  function(palname=pal, ...){
  scale_fill_brewer(palette=palname, ...)
}

knitr::opts_chunk$set(out.width = "225%", out.height = 10)
```



# IDs with Fuso. > .01

## NCI Data

```{r}

# Transform to relative abundance. Save as new object.
ra.dat = transform_sample_counts(phylo.data.nci.umd, function(x){x / sum(x)})
mphyseq = phyloseq::psmelt(ra.dat)
mphyseq <- subset(mphyseq, Abundance > 0.01)

A <- mphyseq %>%
  filter(Genus == "g__Fusobacterium", Abundance > .01)

kable(A[, c("Sample", "Abundance")], style= "html", digits=3) %>%
  kable_styling(full_width = T)


```


```{r}


p <- ggplot(mphyseq, aes(x=Phylum, y=Abundance))+
  geom_violin()+
  geom_jitter(alpha=0.3)
p

```

## TCGA RNAseq

```{r}

# Transform to relative abundance. Save as new object.
ra.dat = transform_sample_counts(phylo.data.tcga.RNAseq, function(x){x / sum(x)})
mphyseq = phyloseq::psmelt(ra.dat)

A <- mphyseq %>%
  filter(Genus == "Fusobacterium", Abundance > .01)

kable(A[, c("Sample", "Abundance")], style= "html", digits=3) %>%
  kable_styling(full_width = T)


```


## TCGA WGS

```{r}

# Transform to relative abundance. Save as new object.
ra.dat = transform_sample_counts(phylo.data.tcga.WGS, function(x){x / sum(x)})
mphyseq = phyloseq::psmelt(ra.dat)

A <- mphyseq %>%
  filter(Genus == "Fusobacterium", Abundance > .01)

kable(A[, c("Sample", "Abundance")], style= "html", digits=3) %>%
  kable_styling(full_width = T)


```




# Addtional Information

Table with sample IDs with highest abundance of the following: 

* Fusobacterium nucleatum
* Streptococcus sanguinis
* Campylobacter concisus

## NCI

### Fusobacterium nucleatum

```{r}

# Transform to relative abundance. Save as new object.
ra.dat = transform_sample_counts(phylo.data.nci.umd, function(x){x / sum(x)})
mphyseq = phyloseq::psmelt(ra.dat)

A <- mphyseq %>%
  filter(OTU == "Fusobacterium_nucleatum", Abundance > 0.001)

kable(A[, c("Sample", "Abundance")], style= "html", digits=3) %>%
  kable_styling(full_width = T)


```

### Streptococcus sanguinis

**NOT UNIQUELY FOUND**

Only found as: ` Streptococcus_dentisani:Streptococcus_infantis:Streptococcus_mitis:Streptococcus_oligofermentans:Streptococcus_oralis:Streptococcus_pneumoniae:Streptococcus_pseudopneumoniae:Streptococcus_sanguinis`

```{r}

A <- mphyseq %>%
  filter(OTU %like% "Streptococcus_sanguinis", Abundance > 0.001)

kable(A[, c("OTU","Sample", "Abundance")], style= "html", digits=3) %>%
  kable_styling(full_width = T)


```

### Campylobacter concisus

**NOT FOUND**

Only found: ` Campylobacter_rectus:Campylobacter_showae`

```{r}

A <- mphyseq %>%
  filter(OTU == "Campylobacter_rectus:Campylobacter_showae", Abundance > 0.001)

kable(A[, c("Sample", "Abundance")], style= "html", digits=3) %>%
  kable_styling(full_width = T)


```

## TCGA RNAseq

### Fusobacterium nucleatum

```{r}

# Transform to relative abundance. Save as new object.
ra.dat = transform_sample_counts(phylo.data.tcga.RNAseq, function(x){x / sum(x)})
mphyseq = phyloseq::psmelt(ra.dat)

A <- mphyseq %>%
  filter(OTU %like% "Fusobacterium nucleatum", Abundance > 0.001)


kable(A[, c("OTU","Sample", "Abundance")], style= "html", digits=3) %>%
  kable_styling(full_width = T)


```

### Streptococcus sanguinis


```{r}

A <- mphyseq %>%
  filter(OTU %like% "Streptococcus sanguinis", Abundance > 0.001)

kable(A[, c("OTU","Sample", "Abundance")], style= "html", digits=3) %>%
  kable_styling(full_width = T)


```

### Campylobacter concisus

```{r}

A <- mphyseq %>%
  filter(OTU %like% "Campylobacter concisus", Abundance > 0.001)

kable(A[, c("OTU","Sample", "Abundance")], style= "html", digits=3) %>%
  kable_styling(full_width = T)


```


## TCGA WGS

### Fusobacterium nucleatum

```{r}

# Transform to relative abundance. Save as new object.
ra.dat = transform_sample_counts(phylo.data.tcga.WGS, function(x){x / sum(x)})
mphyseq = phyloseq::psmelt(ra.dat)

A <- mphyseq %>%
  filter(OTU %like% "Fusobacterium nucleatum", Abundance > 0.001)


kable(A[, c("OTU","Sample", "Abundance")], style= "html", digits=3) %>%
  kable_styling(full_width = T)


```

### Streptococcus sanguinis


```{r}

A <- mphyseq %>%
  filter(OTU %like% "Streptococcus sanguinis", Abundance > 0.001)

kable(A[, c("OTU","Sample", "Abundance")], style= "html", digits=3) %>%
  kable_styling(full_width = T)


```

### Campylobacter concisus

```{r}

A <- mphyseq %>%
  filter(OTU %like% "Campylobacter concisus", Abundance > 0.001)

kable(A[, c("OTU","Sample", "Abundance")], style= "html", digits=3) %>%
  kable_styling(full_width = T)


```


