---
title: "Results for Ordered Logistic Regression Models"
output:
  workflowr::wflow_html:
    toc: true
editor_options:
  chunk_output_type: console
---

# Question 3

```
Q3: Is fuso associated with tumor stage (pTNM) in either data set? Does X bacteria predict stage? Multivariable w/ age, sex, BMI, history of Barrett's
```

Add to this analysis:

* Fusobacterium nucleatum
* Streptococcus sanguinis
* Campylobacter concisus 
* Prevotella spp. 


TCGA drop "not reported" from tumor stage.

# Update as of 2021-03-04

The goal of this analysis is still to answer the question posed above. 
However, the methods by which this is done has been expanded upon and the model has been built to be as interpretable as possible.
In the results, a deeper explanation of the results is given to help explain the effects of adding covariates and how to interpret the parameters.
In general, the model is set up to model how the probability of being in higher tumor stage depends on the individual characteristics of the study participants.
Let $TS$ represents the tumor stage with levels $j = 1, 2, ..., J$ (TS have levels 0, 1, I, II, III, and IV).
We aim to model the odds of being less than or equal to a particular category
\[\frac{Pr(TS\leq j)}{Pr(TS> j)}, j = 1, 2, ..., J-1\]
We aim to model the logit, which is
\[log\left(\frac{Pr(TS\leq j)}{Pr(TS> j)}\right)=logit\left(Pr(TS\leq j)\right)=\beta_{0j}-\left(\eta_1x_1+\eta_2x_2+...+\eta_px_p\right)\]
where the terms on the right represent how R parameterizes the ordered logit model with $\beta_{0j}$ being the intercept/threshold parameter separating category $j$ from $j+1$.
The $\eta_p$ regression weights represent the change in logits.
A major assumption of the ordered logit model, as parameterized here, is the parallel lines assumption.
This means the even though the intercepts differ separate the categories, the regression weights are equal for each successive category.


```{r data, include=FALSE, echo=FALSE, warning=F, error=F, message=F}
# load packages
source("code/load_packages.R")

# get data
source("code/get_cleaned_data.R")

theme_set(theme_bw())
# pal = "Sequential"
# scale_colour_discrete <-  function(palname=pal, ...){
#   scale_colour_brewer(palette=palname, ...)
# }
# scale_fill_discrete <-  function(palname=pal, ...){
#   scale_fill_brewer(palette=palname, ...)
# }

knitr::opts_chunk$set(out.width = "225%", out.height = 10)

options(digits=4)

# transform to relative abundances 
# in scale of 0-100
phylo.data.nci.umd <- transform_sample_counts(phylo.data.nci.umd, function(x){100*(x / sum(x))})
phylo.data.tcga.RNAseq <- transform_sample_counts(phylo.data.tcga.RNAseq, function(x){100*(x / sum(x))})
phylo.data.tcga.WGS <- transform_sample_counts(phylo.data.tcga.WGS, function(x){100*(x / sum(x))})

# melt data down for use
dat.16s <- psmelt(phylo.data.nci.umd)
dat.rna <- psmelt(phylo.data.tcga.RNAseq)
dat.wgs <- psmelt(phylo.data.tcga.WGS)

# fix otu formatting
dat.rna$otu2 <- "a"
dat.wgs$otu2 <- "a"
i <- 1
for(i in 1:nrow(dat.rna)){
  dat.rna$otu2[i] <- str_split(dat.rna$OTU[i], ";")[[1]][7]
}
for(i in 1:nrow(dat.wgs)){
  dat.wgs$otu2[i] <- str_split(dat.wgs$OTU[i], ";")[[1]][7]
}

# set up variables
dat.16s <- dat.16s %>%
  group_by(OTU) %>%
  mutate(
    BarrettsHist = ifelse(Barretts. == "Y", 1, 0),
    bar.c = BarrettsHist - mean(BarrettsHist),
    female = ifelse(gender=="F", 1, 0),
    female.c = female-mean(female),
    age.c = age - mean(age),
    BMI.n = ifelse(BMI == "U", NA, BMI),
    BMI.n = as.numeric(BMI.n),
    bmi.c = BMI.n - mean(BMI.n, na.rm=T),
    bmi.c = ifelse(is.na(bmi.c), 0, bmi.c)
  )

dat.rna <- dat.rna %>%
  #group_by(OTU) %>%
  mutate(
    BarrettsHist = ifelse(Barrett.s.Esophagus.Reported == "Yes", 1, 0),
    BarrettsHist = ifelse(BarrettsHist == "Not Available", NA, BarrettsHist),
    bar.c = BarrettsHist - mean(BarrettsHist),
    female = ifelse(Gender=="female", 1, 0),
    female.c = female-mean(female),
    year_of_death = ifelse(is.na(year_of_death), 2021, year_of_death),
    age = year_of_death-year_of_birth,
    age.c = age - mean(age, na.rm=T),
    bmi.c = bmi - mean(bmi, na.rm=T),
    bmi.c = ifelse(is.na(bmi.c), 0, bmi.c)
  )

dat.wgs <- dat.wgs %>%
  #group_by(OTU) %>%
  mutate(
    BarrettsHist = ifelse(Barrett.s.Esophagus.Reported == "Yes", 1, 0),
    BarrettsHist = ifelse(BarrettsHist == "Not Available", NA, BarrettsHist),
    bar.c = BarrettsHist - mean(BarrettsHist, na.rm=T),
    female = ifelse(Gender=="female", 1, 0),
    female.c = female-mean(female, na.rm=T),
    year_of_death = ifelse(is.na(year_of_death), 2021, year_of_death),
    age = year_of_death-year_of_birth,
    age.c = age - mean(age, na.rm=T),
    bmi.c = bmi - mean(bmi, na.rm=T),
    bmi.c = ifelse(is.na(bmi.c), 0, bmi.c)
  )

dat.16s$sample_type <- 0
dat.16s$sample_type[dat.16s$tissue=="T" &
                      dat.16s$Histology=="ADC" &
                      dat.16s$Barretts. == "Y"] <- "EAC tissues w/ Barretts History"
dat.16s$sample_type[dat.16s$tissue=="N" &
                      dat.16s$Histology=="ADC" &
                      dat.16s$Barretts. == "Y"] <- "EAC-adjacent tissue w/ Barretts History"
dat.16s$sample_type[dat.16s$tissue=="BO" &
                      dat.16s$Histology=="Barrets only"&
                      dat.16s$Barretts. == "Y"] <-"Barretts Only"


dat.rna$sample_type <- 0
dat.rna$sample_type[(dat.rna$morphology=="8140/3" |
                       dat.rna$morphology=="8480/3") &
                       dat.rna$SampleType_Level2=="Tumor" &
                       dat.rna$Barrett.s.Esophagus.Reported=="Yes"] <- "EAC tissues w/ Barretts History"
dat.rna$sample_type[(dat.rna$morphology=="8140/3" | 
                       dat.rna$morphology=="8480/3") &
                       dat.rna$SampleType_Level2=="Normal"&
                       dat.rna$Barrett.s.Esophagus.Reported=="Yes"] <- "EAC-adjacent tissue w/ Barretts History"

dat.rna$EACcomp <- 0
dat.rna$EACcomp[(dat.rna$morphology=="8140/3" |
                       dat.rna$morphology=="8480/3") &
                       dat.rna$SampleType_Level2=="Tumor" &
                       dat.rna$Barrett.s.Esophagus.Reported=="Yes"] <- "EAC tissues w/ Barretts History"
dat.rna$EACcomp[(dat.rna$morphology=="8140/3" |
                       dat.rna$morphology=="8480/3") &
                       dat.rna$SampleType_Level2=="Tumor" &
                       dat.rna$Barrett.s.Esophagus.Reported=="No"] <- "EAC tissues w/o Barretts History"


dat.wgs$sample_type <- 0
dat.wgs$sample_type[(dat.wgs$morphology=="8140/3" |
                       dat.wgs$morphology=="8480/3") &
                       dat.wgs$SampleType_Level2=="Tumor" &
                       dat.wgs$Barrett.s.Esophagus.Reported=="Yes"] <- "EAC tissues w/ Barretts History"
dat.wgs$sample_type[(dat.wgs$morphology=="8140/3" | 
                       dat.wgs$morphology=="8480/3") &
                       dat.wgs$SampleType_Level2=="Normal"&
                       dat.wgs$Barrett.s.Esophagus.Reported=="Yes"] <- "EAC-adjacent tissue w/ Barretts History"

dat.wgs$EACcomp <- 0
dat.wgs$EACcomp[(dat.wgs$morphology=="8140/3" |
                       dat.wgs$morphology=="8480/3") &
                       dat.wgs$SampleType_Level2=="Tumor" &
                       dat.wgs$Barrett.s.Esophagus.Reported=="Yes"] <- "EAC tissues w/ Barretts History"
dat.wgs$EACcomp[(dat.wgs$morphology=="8140/3" |
                       dat.wgs$morphology=="8480/3") &
                       dat.wgs$SampleType_Level2=="Tumor" &
                       dat.wgs$Barrett.s.Esophagus.Reported=="No"] <- "EAC tissues w/o Barretts History"


# make tumor vs normal variable
dat.16s$tumor.cat <- factor(dat.16s$tissue, levels=c("BO", "N", "T"), labels = c("Non-Tumor", "Non-Tumor", "Tumor"))
dat.rna$tumor.cat <- factor(dat.rna$SampleType_Level2, levels=c("Normal", "Tumor"), labels = c("Non-Tumor", "Tumor"))
dat.wgs$tumor.cat <- factor(dat.wgs$SampleType_Level2, levels=c("Normal", "Tumor"), labels = c("Non-Tumor", "Tumor"))

# dataset id
dat.16s$source <- "16s"
dat.rna$source <- "rna"
dat.wgs$source <- "wgs"

# plotting ids
dat.16s$X <- paste0(dat.16s$source, "-", dat.16s$tumor.cat)
dat.rna$X <- paste0(dat.rna$source, "-", dat.rna$tumor.cat)
dat.wgs$X <- paste0(dat.wgs$source, "-", dat.wgs$tumor.cat)

# relabel as (0/1) for analysis
dat.16s$tumor <- as.numeric(factor(dat.16s$tissue, levels=c("BO", "N", "T"), labels = c("Non-Tumor", "Non-Tumor", "Tumor"))) - 1
dat.rna$tumor <- as.numeric(factor(dat.rna$SampleType_Level2, levels=c("Normal", "Tumor"), labels = c("Non-Tumor", "Tumor"))) - 1
dat.wgs$tumor <- as.numeric(factor(dat.wgs$SampleType_Level2, levels=c("Normal", "Tumor"), labels = c("Non-Tumor", "Tumor"))) - 1

# presence- absence
dat.16s$pres <- ifelse(dat.16s$Abundance > 0, 1, 0)
dat.16s$pres[is.na(dat.16s$pres)] <- 0
dat.rna$pres <- ifelse(dat.rna$Abundance > 0, 1, 0)
dat.rna$pres[is.na(dat.rna$pres)] <- 0
dat.wgs$pres <- ifelse(dat.wgs$Abundance > 0, 1, 0)
dat.wgs$pres[is.na(dat.wgs$pres)] <- 0

# tumor stage variable
dat.16s$tumor.stage <- factor(
  dat.16s$pTNM,
  levels = c("0", "1", "I", "IIA", "IIB", "III", "IV"),
  labels = c("0", "1", "I", "II", "II", "III", "IV"), ordered=T
)

dat.rna[dat.rna == "not reported"] <- NA
dat.rna$tumor.stage <- factor(
  dat.rna$tumor_stage,
  levels = c("stage i", "stage ia", "stage ib", "stage iib", "stage iia", "stage iii", "stage iiia", "stage iiib", "stage iiic", "stage iv", "stage iva"),
  labels = c("I", "I", "I", "II", "II", "III", "III", "III", "III", "IV", "IV"), ordered=T
)

dat.wgs[dat.wgs == "not reported"] <- NA
dat.wgs$tumor.stage <- factor(
  dat.wgs$tumor_stage,
  levels = c("stage i", "stage ia", "stage ib", "stage iib", "stage iia", "stage iii", "stage iiia", "stage iiib", "stage iiic", "stage iv", "stage iva"),
  labels = c("I", "I", "I", "II", "II", "III", "III", "III", "III", "IV", "IV"), ordered=T
)


# subset to fuso. nuc. only
# Streptococcus sanguinis 
# Campylobacter concisus
# Prevotella spp.

# Not uniquely identified in 16s
dat.16s.s <- filter(
  dat.16s,
  OTU %in% c(
    "Fusobacterium_nucleatum",
    unique(dat.16s$OTU[dat.16s$OTU %like% "Streptococcus_sanguinis"]),
    unique(dat.16s$OTU[dat.16s$OTU %like% "Campylobacter_"]),
    unique(dat.16s$OTU[dat.16s$OTU %like% "Prevotella"]))
)
dat.rna.s <- filter(
  dat.rna,
  otu2 %in% c(
    "Fusobacterium nucleatum",
    "Streptococcus sanguinis",
    "Campylobacter concisus",
    unique(dat.rna$otu2[dat.rna$otu2 %like% "Prevotella"]))
)
dat.wgs.s <- filter(
  dat.wgs,
  otu2 %in% c(
    "Fusobacterium nucleatum",
    "Streptococcus sanguinis",
    "Campylobacter concisus",
    unique(dat.wgs$otu2[dat.wgs$otu2 %like% "Prevotella"]))
)

# new names -> due to lack of uniqueness
dat.16s.s$OTU1 <- factor(
  dat.16s.s$OTU,
  levels = c(
    "Fusobacterium_nucleatum",
    "Streptococcus_dentisani:Streptococcus_infantis:Streptococcus_mitis:Streptococcus_oligofermentans:Streptococcus_oralis:Streptococcus_pneumoniae:Streptococcus_pseudopneumoniae:Streptococcus_sanguinis",
    "Campylobacter_rectus:Campylobacter_showae",
    "Prevotella_melaninogenica"
  ),
  labels = c(
    "Fusobacterium nucleatum",
    "Streptococcus spp.*",
    "Campylobacter spp.*",
    "Prevotella spp."
  )
)
dat.rna.s$OTU1 <- factor(
  dat.rna.s$otu2,
  levels = c(
    "Fusobacterium nucleatum",
    "Streptococcus sanguinis",
    "Campylobacter concisus",
    "Prevotella melaninogenica",
      "Prevotella denticola",
      "Prevotella intermedia",
      "Prevotella ruminicola"),
  labels = c(
    "Fusobacterium nucleatum",
    "Streptococcus sanguinis",
    "Campylobacter concisus",
    "Prevotella spp.", 
      "Prevotella spp.", 
      "Prevotella spp.", 
      "Prevotella spp.")
)

dat.wgs.s$OTU1 <- factor(
  dat.wgs.s$otu2,
  levels = c(
    "Fusobacterium nucleatum",
    "Streptococcus sanguinis",
    "Campylobacter concisus",
    "Prevotella melaninogenica",
      "Prevotella denticola",
      "Prevotella intermedia",
      "Prevotella ruminicola"),
  labels = c(
    "Fusobacterium nucleatum",
    "Streptococcus sanguinis",
    "Campylobacter concisus",
    "Prevotella spp.", 
      "Prevotella spp.", 
      "Prevotella spp.", 
      "Prevotella spp.")
)

# rename bacteria/OTU variables
dat.16s.s$OTU <- dat.16s.s$OTU1
dat.rna.s$OTU <- dat.rna.s$OTU1
dat.wgs.s$OTU <- dat.wgs.s$OTU1
dat.rna$OTU <- dat.rna$otu2
dat.wgs$OTU <- dat.wgs$otu2

# fix abundance/OTU in rna and WGS data
dat.rna.s <- dat.rna.s %>%
  dplyr::group_by(OTU, ID)%>%
  mutate(Abundance = mean(Abundance))%>%
  distinct(OTU, ID, .keep_all = T)
dat.wgs.s <- dat.wgs.s %>%
  dplyr::group_by(OTU, ID)%>%
  mutate(Abundance = mean(Abundance))%>%
  distinct(OTU, ID, .keep_all = T)
```



# NCI 16s data

## Double Checking Data

```{r nci-fig1, fig.height=5, out.width="100%", out.height="100%"}

# in long format
table(dat.16s$tumor.stage)
# by subject
dat <- dat.16s %>% filter(OTU == "Fusobacterium_nucleatum")
table(dat$tumor.stage)
sum(table(dat$tumor.stage)) # sample size match


mean.dat <- dat.16s.s %>%
  group_by(tumor.stage, OTU) %>%
  summarize(M = mean(Abundance))
ggplot(dat.16s.s, aes(x=tumor.stage, y=Abundance))+
  geom_violin()+
  geom_jitter(alpha=0.25,width = 0.25)+
  geom_point(data=mean.dat, aes(x=tumor.stage, y = M), size=2, alpha =0.9, color="red")+
  labs(x="Tumor Stage", y="Percent Relative Abundance",
       title="Distribution of abundance across tumor stage",
       subtitle="Red dot is average percent relative abundance")+
  scale_y_continuous(trans="pseudo_log")+
  facet_wrap(.~OTU, nrow=1, scales="free")+
  theme_classic()

```

**NOTE** the bacteria were not uniquely defined for 16S data.
That means that these are the bacteria truly used:

* "Fusobacterium_nucleatum"
* Streptococcus spp. = "Streptococcus_dentisani:Streptococcus_infantis:Streptococcus_mitis:Streptococcus_oligofermentans:Streptococcus_oralis:Streptococcus_pneumoniae:Streptococcus_pseudopneumoniae:Streptococcus_sanguinis",
* Campylobacter spp. = "Campylobacter_rectus:Campylobacter_showae",
* Prevotella spp. = "Prevotella_melaninogenica"

Stage "1" has only 1 unique sample and will be dropped from subsequent analyses.
And remove NA values (7).

```{r nci-data-manip1}
dat.16s.s <- dat.16s.s %>%
  filter(tumor.stage != "1")%>%
  mutate(tumor.stage = droplevels(tumor.stage, exclude=c("1",NA)))
```

Now, transform data into a useful wide-format representation to avoid inflating sample size.
The long-format above is very useful for making plots, but less useful for the ordered logistic regression to follow.

```{r nci-data-manip2}

dat.16s.wide <- dat.16s.s %>%
  dplyr::select(OTU, Abundance, Sample.ID, BarrettsHist, bar.c, female, female.c, age.c,BMI.n, bmi.c, sample_type, tumor.cat, tumor.stage) %>%
  pivot_wider(names_from = OTU, values_from = Abundance)

```

## Ordered Logistic Regression

### Model 0: TS ~ 1

This in the intercept only model. Implying that we will only estimate the unique intercepts for the $J-1$ categories.

```{r glm-16s-m0}
## fit ordered logit model and store results 'm'
# use only 1 bacteria
fit <- fit0 <- MASS::polr(tumor.stage ~ 1, data = filter(dat.16s.s, OTU =="Fusobacterium nucleatum"), Hess=TRUE)
## view a summary of the model
summary(fit)
# save fitted probabilities
# these represent the estimated probability of each tumor stage
pp <- fitted(fit)
pp[1,]
```

The model can be more fully described as
\begin{align*}
logit\left(Pr(TS\leq 0)\right)&=-0.95\\
logit\left(Pr(TS\leq I)\right)&=-0.24\\
logit\left(Pr(TS\leq II)\right)&=1.27\\
logit\left(Pr(TS\leq III)\right)&=2.64\\
\end{align*}

Interpreting the intercepts only model is relatively straightforward. The main interpretation we can obtain from this base model is the probability of each category.
We obtain these probabilities by applying the inverse-logit transformation to obtain each of the cumulative probabilities and then use the relevant probabilities to compute each individual category probability.
That is, 
\begin{align*}
Pr(TS = 0)&= Pr(TS\leq 0) = 0.28\\
Pr(TS = I)&=Pr(TS\leq I) - Pr(TS\leq 0)= 0.44-0.28 =0.16\\
Pr(TS = II)&=Pr(TS\leq II) - Pr(TS\leq I) = 0.78 - 0.44 = 0.34 \\
Pr(TS = III)&=Pr(TS\leq III) - Pr(TS\leq II) =0.93-0.78=0.15\\
Pr(TS = IV)&=1 - Pr(TS\leq IV) =1-0.93=0.07,\\
\end{align*}
which, in all is not very informative.
More information is definately gained as predictors/covariates are introduced.

### Model 1: TS ~ OTU

This model expands on the previous model to include OTU abundance as a predictor of TS.
OTU abundance is captured by four variables as predictors of TS (the four OTUs listed at the top of this document).


```{r glm-16s-m1, fig.height=5, out.width="100%", out.height="100%"}

dat0 <- dat.16s.wide
## fit ordered logit model and store results 'm'
fit <- fit1 <- MASS::polr(tumor.stage ~ 1 + `Fusobacterium nucleatum` + `Streptococcus spp.*` + `Campylobacter spp.*` + `Prevotella spp.`, data = dat.16s.wide, Hess=TRUE)

## view a summary of the model
summary(fit)
anova(fit1, fit0) # Chi-square difference test
# obtain approximate p-values
ctable <- coef(summary(fit))
p <- pnorm(abs(ctable[, "t value"]), lower.tail = FALSE) * 2
(ctable <- cbind(ctable, "p value" = p))

# obtain CIs
(ci <- confint(fit))  # CIs assuming normality

## OR and CI
exp(cbind(OR = coef(fit), ci))

# save fitted logits
pp <- fitted(fit)
# predictive data
dat0 <- cbind(dat0, predict(fit, newdata = dat.16s.wide, "probs"))

## melt data set to long for ggplot2
dat1 <- dat0 %>%
  pivot_longer(
    cols=`0`:`IV`,
    names_to = "TS",
    values_to = "Pred.Prob"
  ) %>%
  pivot_longer(
    cols=`Streptococcus spp.*`:`Campylobacter spp.*`,
    names_to="OTU",
    values_to="Abundance"
  )
## plot predicted probabilities across Abundance values for each level of OTU
## facetted by tumor.stage
ggplot(dat1, aes(x = Abundance, y = Pred.Prob, color=TS, group=TS, linetype=TS)) +
  geom_line() + 
  facet_grid(OTU ~., scales="free")+
  labs(y="Probability of Tumor Stage",
       x="Relative Abundance (%)",
       title="Tumor stage likelihood with bacteria abundance",
       color="Tumor Stage", linetype="Tumor Stage")+
  theme(
    panel.grid = element_blank()
  )
```

The results from the model when the abundance of the four OTUs are included in the model becomes a bit more complicated.
First of all, the model with OTU abundances did not fit better in terms of AIC, BIC, and no difference was found from a chi-square difference test.

For interpretation purposes, the following is how to interpret the effects of OTU abundance on predicting tumor stage.
First, the logit model is
\begin{align*}
logit\left(Pr(TS\leq 0)\right)&=-0.95\\
logit\left(Pr(TS\leq I)\right)&=-0.24\\
logit\left(Pr(TS\leq II)\right)&=1.27\\
logit\left(Pr(TS\leq III)\right)&=2.64\\
\end{align*}
While interpreting the odds ratios can be done as follows.
First, take Streptococcus spp., the OR was 1.00 95% CI (0.99, 1.01).
This implies that for individuals who had a 1% higher relative abundance of Streptococcus spp., the odds of being in a higher tumor stage is multiplied 1.00 times (i.e., 0% increase on average) holding the abundance of other OTUs constant.
This means that the knowledge of relative abundance of Strepto was not informative over the base probability we estimated in model 0.

For Fusobacterium nucleatum (OR = 1.01, 95% CI [0.99, 1.04]) and Prevotella spp. (OR = 1.01, 95% CI [0.99, 1.05]), the interpretation is as follows because the OR is the same.
For individuals who had a 1% higher relative abundance of this OTU (Fuso. or Prevo.), the odds of being in a higher tumor stage is multiplied 1.01 times (i.e., about a 1% increase on average) holding the abundance of other OTUs constant.

### Model 2: TS ~ OTU + COVARIATES

```{r glm-16s-m2, fig.height=5, out.width="100%", out.height="100%"}

dat0 <- dat.16s.wide
## fit ordered logit model and store results 'm'
fit <- fit2 <- MASS::polr(tumor.stage ~ 1+ `Fusobacterium nucleatum` + `Streptococcus spp.*` + `Campylobacter spp.*` + `Prevotella spp.` + age.c + female.c + bmi.c + bar.c, data = dat.16s.wide, Hess=TRUE)

## view a summary of the model
summary(fit)
anova(fit2, fit0) # Chi-square difference test
# obtain approximate p-values
ctable <- coef(summary(fit))
p <- pnorm(abs(ctable[, "t value"]), lower.tail = FALSE) * 2
(ctable <- cbind(ctable, "p value" = p))

# obtain CIs
(ci <- confint(fit))  # CIs assuming normality

## OR and CI
exp(cbind(OR = coef(fit), ci))

# save fitted logits
pp <- fitted(fit)
# predictive data
dat0 <- cbind(dat0, predict(fit, newdata = dat.16s.wide, "probs"))

## melt data set to long for ggplot2
dat1 <- dat0 %>%
  pivot_longer(
    cols=`0`:`IV`,
    names_to = "TS",
    values_to = "Pred.Prob"
  ) %>%
  pivot_longer(
    cols=`Streptococcus spp.*`:`Campylobacter spp.*`,
    names_to="OTU",
    values_to="Abundance"
  )
## plot predicted probabilities across Abundance values for each level of OTU
## facetted by tumor.stage
ggplot(dat1, aes(x = Abundance, y = Pred.Prob, color=TS, group=TS, linetype=TS)) +
  geom_line() + 
  facet_grid(OTU ~., scales="free")+
  labs(y="Probability of Tumor Stage",
       x="Relative Abundance (%)",
       title="Tumor stage likelihood with bacteria abundance",
       color="Tumor Stage", linetype="Tumor Stage")+
  theme(
    panel.grid = element_blank()
  )
```


No significant changes. Everything is still insignificant.
For interpretation purposes, the results in from this model can be interpreted the same as the previous because all covariates were mean-centered.

### Proportional Odds Assumption

```{r pro-odds-test1}
dat.16s.wide <- dat.16s.wide %>%
  mutate(
    T0 = I(as.numeric(tumor.stage) >= 1),
    TI = I(as.numeric(tumor.stage) >= 2),
    TII = I(as.numeric(tumor.stage) >= 3),
    TIII = I(as.numeric(tumor.stage) >= 4),
    TIV = I(as.numeric(tumor.stage) >= 5),
  )
apply(dat.16s.wide[,c('TI','TII','TIII', 'TIV')],2, table)

summary(glm(TI  ~ 1 + `Fusobacterium nucleatum` + `Streptococcus spp.*` + `Campylobacter spp.*` + `Prevotella spp.`+ age.c + female.c + bmi.c + bar.c, data = dat.16s.wide, family=binomial(link="logit")))

summary(glm(TII ~ 1 + `Fusobacterium nucleatum` + `Streptococcus spp.*` + `Campylobacter spp.*` + `Prevotella spp.`+ age.c + female.c + bmi.c + bar.c, data = dat.16s.wide, family=binomial(link="logit")))

summary(glm(TIII ~ 1 + `Fusobacterium nucleatum` + `Streptococcus spp.*` + `Campylobacter spp.*` + `Prevotella spp.`+ age.c + female.c + bmi.c + bar.c, data = dat.16s.wide, family=binomial(link="logit")))

summary(glm(TIV  ~ 1 + `Fusobacterium nucleatum` + `Streptococcus spp.*` + `Campylobacter spp.*` + `Prevotella spp.` + age.c + female.c + bmi.c + bar.c, data = dat.16s.wide, family=binomial(link="logit")))

```

# TCGA RNAseq data

## Double Checking Data

```{r rna-fig1, fig.height=5, out.width="100%", out.height="100%"}

# in long format
table(dat.rna.s$tumor.stage)
# by subject
dat <- dat.rna.s %>% filter(OTU == "Fusobacterium nucleatum")
table(dat$tumor.stage)
sum(table(dat$tumor.stage)) 
nrow(dat) # matches but there's missing data
sum(is.na(dat$Abundance))
# number of non-missing abundnance data
nrow(dat) - sum(is.na(dat$Abundance))
# matches 
dat0 <- dat %>%
  filter()
table(dat0$tumor.stage)

# plot functions
#root function
root<-function(x){
  x <- ifelse(x < 0, 0, x)
  x**(0.2)
}
#inverse root function
invroot<-function(x){
  x**(5)
}

mean.dat <- dat.rna.s %>%
  group_by(tumor.stage, OTU) %>%
  summarize(M = mean(Abundance, na.rm=T))
ggplot(dat.rna.s, aes(x=tumor.stage, y=Abundance))+
  geom_violin()+
  geom_jitter(alpha=0.25,width = 0.25)+
  geom_point(data=mean.dat, aes(x=tumor.stage, y = M), size=2, alpha =0.9, color="red")+
  labs(x="Tumor Stage", y="Percent Relative Abundance",
       title="Distribution of abundance across tumor stage",
       subtitle="Red dot is average percent relative abundance")+
  scale_y_continuous(
    trans=scales::trans_new("root", root, invroot),
    breaks=c(0, 0.001,0.01, 0.1, 1,10,50),
    labels = c(0, 0.001,0.01, 0.1, 1,10,50),
    limits = c(0, 50)
  )+
  facet_wrap(.~OTU, nrow=1, scales="free")+
  theme_classic()

```


Stage IV only has two cases, remove for analysis.

```{r rna-data-manip1}
dat.rna.s <- dat.rna.s %>%
  filter(is.na(Abundance) == F, tumor.stage!="IV")%>%
  mutate(
    tumor.stage = factor(
      tumor.stage,
      levels = c("I", "II", "III"), ordered=T
    )
  )
  
```

Now, transform data into a useful wide-format representation to avoid inflating sample size.
The long-format above is very useful for making plots, but less useful for the ordered logistic regression to follow.

```{r rna-data-manip2}

dat.rna.wide <- dat.rna.s %>%
  dplyr::select(OTU, Abundance, ID, BarrettsHist, bar.c, female.c, age.c, bmi.c, tumor.stage) %>%
  
  pivot_wider(names_from = OTU, values_from = Abundance)
dat.rna.wide$age.c[is.na(dat.rna.wide$age.c)]<-0
```

## Ordered Logistic Regression

### Model 0: TS ~ 1

This in the intercept only model. Implying that we will only estimate the unique intercepts for the $J-1$ categories.

```{r glm-rna-m0}
## fit ordered logit model and store results 'm'
fit <- fit0 <- MASS::polr(tumor.stage ~ 1, data = dat.rna.wide, Hess=TRUE)
## view a summary of the model
summary(fit)
# save fitted probabilities
# these represent the estimated probability of each tumor stage
pp <- fitted(fit)
pp[1,]
```



### Model 1: TS ~ OTU

This model expands on the previous model to include OTU abundance as a predictor of TS.
OTU abundance is captured by four variables as predictors of TS (the four OTUs listed at the top of this document).


```{r glm-rna-m1, fig.height=5, out.width="100%", out.height="100%"}

dat0 <- dat.rna.wide
## fit ordered logit model and store results 'm'
fit <- fit1 <- MASS::polr(tumor.stage ~ 1 + `Fusobacterium nucleatum` + `Streptococcus sanguinis` + `Campylobacter concisus` + `Prevotella spp.`, data = dat.rna.wide, Hess=TRUE)

## view a summary of the model
summary(fit)

# rescale variables
dat.rna.wide <- dat.rna.wide %>%
  mutate(
    `Fusobacterium nucleatum` = 10*`Fusobacterium nucleatum`,
    `Streptococcus sanguinis` = 10*`Streptococcus sanguinis`,
    `Campylobacter concisus` = 10*`Campylobacter concisus`,
    `Prevotella spp.` = 10*`Prevotella spp.`
  )
dat0 <- dat.rna.wide
fit <- fit1 <- MASS::polr(tumor.stage ~ 1 + `Fusobacterium nucleatum` + `Streptococcus sanguinis` + `Campylobacter concisus` + `Prevotella spp.`, data = dat.rna.wide, Hess=TRUE)
# interpret in terms of change in 0.1% increase in relative abundance

## view a summary of the model
summary(fit)

anova(fit1, fit0) # Chi-square difference test
# obtain approximate p-values
ctable <- coef(summary(fit))
p <- pnorm(abs(ctable[, "t value"]), lower.tail = FALSE) * 2
(ctable <- cbind(ctable, "p value" = p))

# obtain CIs
(ci <- confint(fit))  # CIs assuming normality

## OR and CI
exp(cbind(OR = coef(fit), ci))

# predictive data
pp <- as.data.frame(predict(fit, newdata = dat.rna.wide, "probs"))
dat0 <- cbind(dat0, pp)

## melt data set to long for ggplot2
dat1 <- dat0 %>%
  pivot_longer(
    cols=`I`:`III`,
    names_to = "TS",
    values_to = "Pred.Prob"
  ) %>%
  pivot_longer(
    cols=`Fusobacterium nucleatum`:`Campylobacter concisus`,
    names_to="OTU",
    values_to="Abundance"
  )
## plot predicted probabilities across Abundance values for each level of OTU
## facetted by tumor.stage
ggplot(dat1, aes(x = Abundance, y = Pred.Prob, color=TS, group=TS, linetype=TS)) +
  geom_line() + 
  facet_grid(OTU ~., scales="free")+
  labs(y="Probability of Tumor Stage",
       x="Relative Abundance (% times 10)",
       title="Tumor stage likelihood with bacteria abundance",
       color="Tumor Stage", linetype="Tumor Stage")+
  scale_x_continuous(
    trans=scales::trans_new("root", root, invroot),
    breaks=c(0, 0.001,0.01, 0.1, 1,10,25),
    labels = c(0, 0.001,0.01, 0.1, 1,10,25),
    limits = c(0, 25)
  )+
  theme(
    panel.grid = element_blank()
  )
```



### Model 2: TS ~ OTU + COVARIATES

```{r glm-rna-m2, fig.height=5, out.width="100%", out.height="100%"}

dat0 <- dat.rna.wide
## fit ordered logit model and store results 'm'
fit <- fit2 <- MASS::polr(tumor.stage ~ 1+ `Fusobacterium nucleatum` + `Streptococcus sanguinis` + `Campylobacter concisus` + `Prevotella spp.` + age.c + female.c + bmi.c + bar.c, data = dat.rna.wide, Hess=TRUE)

## view a summary of the model
summary(fit)
anova(fit2, fit0) # Chi-square difference test
# obtain approximate p-values
ctable <- coef(summary(fit))
p <- pnorm(abs(ctable[, "t value"]), lower.tail = FALSE) * 2
(ctable <- cbind(ctable, "p value" = p))

# obtain CIs
(ci <- confint(fit))  # CIs assuming normality

## OR and CI
exp(cbind(OR = coef(fit), ci))

# predictive data
pp <- as.data.frame(predict(fit, newdata = dat.rna.wide, "probs"))
dat0 <- cbind(dat0, pp)

## melt data set to long for ggplot2
dat1 <- dat0 %>%
  pivot_longer(
    cols=`I`:`III`,
    names_to = "TS",
    values_to = "Pred.Prob"
  ) %>%
  pivot_longer(
    cols=`Fusobacterium nucleatum`:`Campylobacter concisus`,
    names_to="OTU",
    values_to="Abundance"
  )
## plot predicted probabilities across Abundance values for each level of OTU
## facetted by tumor.stage
ggplot(dat1, aes(x = Abundance, y = Pred.Prob, color=TS, group=TS, linetype=TS)) +
  geom_line() + 
  facet_grid(OTU ~., scales="free")+
  labs(y="Probability of Tumor Stage",
       x="Relative Abundance (% time 10)",
       title="Tumor stage likelihood with bacteria abundance",
       color="Tumor Stage", linetype="Tumor Stage")+
  scale_x_continuous(
    trans=scales::trans_new("root", root, invroot),
    breaks=c(0, 0.001,0.01, 0.1, 1,10,25),
    labels = c(0, 0.001,0.01, 0.1, 1,10,25),
    limits = c(0, 25)
  )+
  theme(
    panel.grid = element_blank()
  )
```


### Proportional Odds Assumption

```{r pro-odds-test2}
dat.rna.wide <- dat.rna.wide %>%
  mutate(
    TI = I(as.numeric(tumor.stage) >= 1),
    TII = I(as.numeric(tumor.stage) >= 2),
    TIII = I(as.numeric(tumor.stage) >= 3)
  )
apply(dat.rna.wide[,c('TII','TIII')],2, table)

summary(glm(TII ~ 1 + `Fusobacterium nucleatum` + `Streptococcus sanguinis` + `Campylobacter concisus` + `Prevotella spp.`+ age.c + female.c + bmi.c + bar.c, data = dat.rna.wide, family=binomial(link="logit")))

summary(glm(TIII ~ 1 + `Fusobacterium nucleatum` + `Streptococcus sanguinis` + `Campylobacter concisus` + `Prevotella spp.`+ age.c + female.c + bmi.c + bar.c, data = dat.rna.wide, family=binomial(link="logit")))

```


# TCGA WGS data

## Double Checking Data

```{r wgs-fig1, fig.height=5, out.width="100%", out.height="100%"}

# in long format
table(dat.wgs.s$tumor.stage)
# by subject
dat <- dat.wgs.s %>% filter(OTU == "Fusobacterium nucleatum")
table(dat$tumor.stage)
sum(table(dat$tumor.stage)) 
nrow(dat) # matches but there's missing data
sum(is.na(dat$Abundance))
# number of non-missing abundnance data
nrow(dat) - sum(is.na(dat$Abundance))
# matches 
dat0 <- dat %>%
  filter()
table(dat0$tumor.stage)

# plot functions
#root function
root<-function(x){
  x <- ifelse(x < 0, 0, x)
  x**(0.2)
}
#inverse root function
invroot<-function(x){
  x**(5)
}

mean.dat <- dat.wgs.s %>%
  group_by(tumor.stage, OTU) %>%
  summarize(M = mean(Abundance, na.rm=T))
ggplot(dat.wgs.s, aes(x=tumor.stage, y=Abundance))+
  geom_violin()+
  geom_jitter(alpha=0.25,width = 0.25)+
  geom_point(data=mean.dat, aes(x=tumor.stage, y = M), size=2, alpha =0.9, color="red")+
  labs(x="Tumor Stage", y="Percent Relative Abundance",
       title="Distribution of abundance across tumor stage",
       subtitle="Red dot is average percent relative abundance")+
  scale_y_continuous(
    trans=scales::trans_new("root", root, invroot),
    breaks=c(0, 0.001,0.01, 0.1, 1,10,50),
    labels = c(0, 0.001,0.01, 0.1, 1,10,50),
    limits = c(0, 50)
  )+
  facet_wrap(.~OTU, nrow=1, scales="free")+
  theme_classic()

```


```{r wgs-data-manip1}
dat.wgs.s <- dat.wgs.s %>%
  filter(is.na(Abundance) == F, is.na(tumor.stage)==F)
  
```

Now, transform data into a useful wide-format representation to avoid inflating sample size.
The long-format above is very useful for making plots, but less useful for the ordered logistic regression to follow.

```{r wgs-data-manip2}

dat.wgs.wide <- dat.wgs.s %>%
  dplyr::select(OTU, Abundance, ID, BarrettsHist, bar.c, female.c, age.c, bmi.c, tumor.stage) %>%
  
  pivot_wider(names_from = OTU, values_from = Abundance)

```

## Ordered Logistic Regression

### Model 0: TS ~ 1

This in the intercept only model. Implying that we will only estimate the unique intercepts for the $J-1$ categories.

```{r glm-wgs-m0}
## fit ordered logit model and store results 'm'
fit <- fit0 <- MASS::polr(tumor.stage ~ 1, data = dat.wgs.wide, Hess=TRUE)
## view a summary of the model
summary(fit)
# save fitted probabilities
# these represent the estimated probability of each tumor stage
pp <- fitted(fit)
pp[1,]
```



### Model 1: TS ~ OTU

This model expands on the previous model to include OTU abundance as a predictor of TS.
OTU abundance is captured by four variables as predictors of TS (the four OTUs listed at the top of this document).


```{r glm-wgs-m1, fig.height=5, out.width="100%", out.height="100%"}

dat0 <- dat.wgs.wide
## fit ordered logit model and store results 'm'
fit <- fit1 <- MASS::polr(tumor.stage ~ 1+ `Fusobacterium nucleatum` + `Streptococcus sanguinis` + `Campylobacter concisus` + `Prevotella spp.`, data = dat.wgs.wide, Hess=TRUE)

## view a summary of the model
summary(fit)
anova(fit1, fit0) # Chi-square difference test
# obtain approximate p-values
ctable <- coef(summary(fit))
p <- pnorm(abs(ctable[, "t value"]), lower.tail = FALSE) * 2
(ctable <- cbind(ctable, "p value" = p))

# obtain CIs
(ci <- confint(fit))  # CIs assuming normality

## OR and CI
exp(cbind(OR = coef(fit), ci))

# predictive data
pp <- as.data.frame(predict(fit, newdata = dat.wgs.wide, "probs"))
dat0 <- cbind(dat0, pp)

## melt data set to long for ggplot2
dat1 <- dat0 %>%
  pivot_longer(
    cols=`I`:`IV`,
    names_to = "TS",
    values_to = "Pred.Prob"
  ) %>%
  pivot_longer(
    cols=`Prevotella spp.`:`Streptococcus sanguinis` ,
    names_to="OTU",
    values_to="Abundance"
  )
## plot predicted probabilities across Abundance values for each level of OTU
## facetted by tumor.stage
ggplot(dat1, aes(x = Abundance, y = Pred.Prob, color=TS, group=TS, linetype=TS)) +
  geom_line() + 
  facet_grid(OTU ~., scales="free")+
  labs(y="Probability of Tumor Stage",
       x="Relative Abundance (%)",
       title="Tumor stage likelihood with bacteria abundance",
       color="Tumor Stage", linetype="Tumor Stage")+
  scale_x_continuous(
    trans=scales::trans_new("root", root, invroot),
    breaks=c(0, 0.001,0.01, 0.1, 1,10,50),
    labels = c(0, 0.001,0.01, 0.1, 1,10,50),
    limits = c(0, 50)
  )+
  theme(
    panel.grid = element_blank()
  )
```



### Model 2: TS ~ OTU + COVARIATES

```{r glm-wgs-m2, fig.height=5, out.width="100%", out.height="100%"}

dat0 <- dat.wgs.wide
## fit ordered logit model and store results 'm'
fit <- fit2 <- MASS::polr(tumor.stage ~ 1+ `Fusobacterium nucleatum` + `Streptococcus sanguinis` + `Campylobacter concisus` + `Prevotella spp.` + age.c + female.c + bmi.c + bar.c, data = dat.wgs.wide, Hess=TRUE)

## view a summary of the model
summary(fit)
anova(fit2, fit0) # Chi-square difference test
# obtain approximate p-values
ctable <- coef(summary(fit))
p <- pnorm(abs(ctable[, "t value"]), lower.tail = FALSE) * 2
(ctable <- cbind(ctable, "p value" = p))

# obtain CIs
(ci <- confint(fit))  # CIs assuming normality

## OR and CI
exp(cbind(OR = coef(fit), ci))

# predictive data
pp <- as.data.frame(predict(fit, newdata = dat.wgs.wide, "probs"))
dat0 <- cbind(dat0, pp)

## melt data set to long for ggplot2
dat1 <- dat0 %>%
  pivot_longer(
    cols=`I`:`IV`,
    names_to = "TS",
    values_to = "Pred.Prob"
  ) %>%
  pivot_longer(
    cols=`Prevotella spp.`:`Streptococcus sanguinis` ,
    names_to="OTU",
    values_to="Abundance"
  )
## plot predicted probabilities across Abundance values for each level of OTU
## facetted by tumor.stage
ggplot(dat1, aes(x = Abundance, y = Pred.Prob, color=TS, group=TS, linetype=TS)) +
  geom_line() + 
  facet_grid(OTU ~., scales="free")+
  labs(y="Probability of Tumor Stage",
       x="Relative Abundance (%)",
       title="Tumor stage likelihood with bacteria abundance",
       color="Tumor Stage", linetype="Tumor Stage")+
  scale_x_continuous(
    trans=scales::trans_new("root", root, invroot),
    breaks=c(0, 0.001,0.01, 0.1, 1,10,25),
    labels = c(0, 0.001,0.01, 0.1, 1,10,25),
    limits = c(0, 25)
  )+
  theme(
    panel.grid = element_blank()
  )
```


### Proportional Odds Assumption

```{r pro-odds-test3}
dat.wgs.wide <- dat.wgs.wide %>%
  mutate(
    TI = I(as.numeric(tumor.stage) >= 1),
    TII = I(as.numeric(tumor.stage) >= 2),
    TIII = I(as.numeric(tumor.stage) >= 3),
    TIV = I(as.numeric(tumor.stage) >= 4)
  )
apply(dat.wgs.wide[,c('TII','TIII', "TIV")],2, table)

summary(glm(TII ~ 1 + `Fusobacterium nucleatum` + `Streptococcus sanguinis` + `Campylobacter concisus` + `Prevotella spp.`+ age.c + female.c + bmi.c + bar.c, data = dat.wgs.wide, family=binomial(link="logit")))

summary(glm(TIII ~ 1 + `Fusobacterium nucleatum` + `Streptococcus sanguinis` + `Campylobacter concisus` + `Prevotella spp.`+ age.c + female.c + bmi.c + bar.c, data = dat.wgs.wide, family=binomial(link="logit")))

summary(glm(TIV ~ 1 + `Fusobacterium nucleatum` + `Streptococcus sanguinis` + `Campylobacter concisus` + `Prevotella spp.`+ age.c + female.c + bmi.c + bar.c, data = dat.wgs.wide, family=binomial(link="logit")))


```
