---
title: "PICRUSt Stratified Analyses"
subtitle: "Streptococcus sanguinis"
output:
  workflowr::wflow_html:
    toc: true
editor_options:
  chunk_output_type: console
---

```
i. Mean relative abundance in Tumors

ii. Standard error of mean in Tumors

iii. Mean relative abundance in Non-tumors

iv. Standard error of mean in Non-tumors

v. GLM differential abundance analysis between Tumor vs Non-tumor (Patient=Fixed effect)
  1. P-value
  2. FDR corrected q-value
vi. GLMM differential abundance analysis between Tumor vs Non-tumor (Patient=Random Effect)
  1.P-value
  2. FDR corrected q-value
```

```{r data, include=FALSE, echo=FALSE, warning=F, error=F, message=F}
# load packages
source("code/load_packages.R")

# get data
source("code/get_cleaned_data.R")

theme_set(theme_bw())
# pal = "Sequential"
# scale_colour_discrete <-  function(palname=pal, ...){
#   scale_colour_brewer(palette=palname, ...)
# }
# scale_fill_discrete <-  function(palname=pal, ...){
#   scale_fill_brewer(palette=palname, ...)
# }

knitr::opts_chunk$set(out.width = "225%", out.height = 10)

options(digits=5, scipen=5)

# transform to relative abundances 
# in scale of 0-100
phylo.data.nci.umd <- transform_sample_counts(phylo.data.nci.umd, function(x){100*(x / sum(x))})

# melt data down for use
dat.16s <- phyloseq::sample_data(phylo.data.nci.umd)

# set up variables
dat.16s$sample_type <- 0
dat.16s$sample_type[dat.16s$tissue=="T" &
                      dat.16s$Histology=="ADC" &
                      dat.16s$Barretts. == "Y"] <- "EAC tissues w/ Barretts History"
dat.16s$sample_type[dat.16s$tissue=="N" &
                      dat.16s$Histology=="ADC" &
                      dat.16s$Barretts. == "Y"] <- "EAC-adjacent tissue w/ Barretts History"
dat.16s$sample_type[dat.16s$tissue=="BO" &
                      dat.16s$Histology=="Barrets only"&
                      dat.16s$Barretts. == "Y"] <-"Barretts Only"


# make tumor vs normal variable
dat.16s$tumor.cat <- factor(dat.16s$tissue, levels=c("BO", "N", "T"), labels = c("Non-Tumor", "Non-Tumor", "Tumor"))

# relabel as (0/1) for analysis
dat.16s$tumor <- as.numeric(factor(dat.16s$tissue, levels=c("BO", "N", "T"), labels = c("Non-Tumor", "Non-Tumor", "Tumor"))) - 1

# tumor stage variable
dat.16s$tumor.stage <- factor(
  dat.16s$pTNM,
  levels = c("0", "1", "I", "IIA", "IIB", "III", "IV"),
  labels = c("0", "1", "I", "II", "II", "III", "IV")
)



```


# Part 1 EC Data

```{r pi-dat-ec}

# note: must rename function to func (or something else) because "function" is a reserved name in R and can't be used.

descriptions <- readr::read_tsv("data/PICRUST/ec_pred_metagenome_unstrat_descrip.tsv")[,1:2] %>%
  distinct()
colnames(descriptions) <- c("func", "description")

pi.dat <- readxl::read_xlsx("data/PICRUST-Stratified/EC_strepto.xlsx")
colnames(pi.dat) <- c("func",  colnames(pi.dat)[-c(1)])

# add descriptions to dataframe
pi.dat <- left_join(pi.dat, descriptions, by="func")
pi.dat <- pi.dat[,c(1,161,3:160)]

# aggregate descriptions to get 1 description per row - unique;y defined.
pi.dat <- pi.dat %>%
  group_by(description) %>%
  summarise(across(`1.S37.Jun172016`:`99.D01.S37.Jun232016`,.fns = sum))

# make long format
pi.dat <- pi.dat %>%
  pivot_longer(
    cols=`1.S37.Jun172016`:`99.D01.S37.Jun232016`,
    names_to = "ID",
    values_to = "Abundance"
  )
d <- as.data.frame(dat.16s)
mydata <- full_join(pi.dat, d)

mydata <- mydata %>%
  mutate(
    ID.n = as.numeric(as.factor(ID)),
    description.n = as.numeric(as.factor(description))
  ) %>%
  group_by(ID) %>%
  mutate(
    dem = ifelse(sum(Abundance)==0, 1, sum(Abundance)),
    RelAbundance = Abundance/dem*100) %>%
  ungroup()%>%
  group_by(description)%>%
  mutate(avgRA = mean(RelAbundance))

```


## Abundance

I needed to change from relative abundance to abundance due to so many individuals having 0 counts for all descriptions.

```{r ec-ra}


d <- mydata %>%
  filter(description=="DNA-directed DNA polymerase")

t.test(d$RelAbundance ~ d$tumor.cat)

# Run on all descriptions
tb.ra1 <- mydata %>%
  group_by(description) %>%
  summarise(ng = n(),
            Overall.M = mean(RelAbundance),
            Overall.SE = sd(RelAbundance)/sqrt(ng))
tb.ra2m <- mydata %>%
  group_by(description, tumor.cat) %>%
  summarise(M = mean(RelAbundance)) %>%
  pivot_wider(id_cols = description,
              names_from = tumor.cat,
              values_from = M)
tb.ra2se <- mydata %>%
  group_by(description, tumor.cat) %>%
  summarise(ng=n(),SE = sd(RelAbundance)/sqrt(ng)) %>%
  pivot_wider(id_cols = description,
              names_from = tumor.cat,
              values_from = SE)
tb.ra2var <- mydata %>%
  group_by(description, tumor.cat) %>%
  summarise(ng=n(), VAR = var(RelAbundance)) %>%
  pivot_wider(id_cols = description,
              names_from = tumor.cat,
              values_from = VAR)
tb.ra2ng <- mydata %>%
  group_by(description, tumor.cat) %>%
  summarise(ng=n()) %>%
  pivot_wider(id_cols = description,
              names_from = tumor.cat,
              values_from = ng)
tb.ra <- left_join(tb.ra1, tb.ra2m)
tb.ra <- cbind(tb.ra, tb.ra2se[,-1])
tb.ra <- cbind(tb.ra, tb.ra2var[,-1])
tb.ra <- cbind(tb.ra, tb.ra2ng[,-1]) 

colnames(tb.ra) <- c("description", "ng", "Overall Mean", "Overall SE", "Non-Tumor Mean", "Tumor Mean", "Non-Tumor SE", "Tumor SE","Non-Tumor Var", "Tumor Var", "Non-Tumor Ng", "Tumor Ng")
tb.ra <- tb.ra %>%
  arrange(desc(`Overall Mean`))
tb.ra <- tb.ra[, c("description", "Overall Mean", "Overall SE","Tumor Mean","Tumor Var", "Tumor SE","Tumor Ng", "Non-Tumor Mean","Non-Tumor Var", "Non-Tumor SE", "Non-Tumor Ng")]

# compute t-test
tb.ra <- tb.ra %>%
  mutate(
    SEpooled = sqrt(`Tumor Var`/`Tumor Ng` + `Non-Tumor Var`/`Non-Tumor Ng`),
    t = (`Tumor Mean` - `Non-Tumor Mean`)/(SEpooled),
    df = ((`Tumor Var`/`Tumor Ng` + `Non-Tumor Var`/`Non-Tumor Ng`)**2)/(((`Tumor Var`/`Tumor Ng`)**2)/(`Tumor Ng`-1) + ((`Non-Tumor Var`/`Non-Tumor Ng`)**2)/(`Non-Tumor Ng`-1)),
    p = pt(q = abs(t), df=df, lower.tail = F)*2,
    fdr_p = p.adjust(p, method="fdr")
  )


kable(tb.ra, format="html", digits=5, caption="Stratefied EC Data - Streptococcus sanguinis : Average RelAbundance of Each Description (sorted in descending order)") %>%
  kable_styling(full_width = T) %>%
  scroll_box(width = "100%", height="600px")

```


```{r ec-full-plot, warning=F, error=F, message=F, fig.dim=c(12,10), out.height="100%", out.width="90%"}


plot.dat <- tb.ra %>%
  arrange(desc(`Overall Mean`)) %>%
  slice_head(n=50)%>%
  mutate(
    description = fct_reorder(description, `Overall Mean`),
    description = factor(description, levels = levels(description), ordered=T)
  ) %>%
  arrange(description) %>%
  mutate(
    id = 1:n(),
    step = ifelse(id%%2 == 0, 1, 0),
    Mean_diff = `Tumor Mean` - `Non-Tumor Mean`,
    diff_se = SEpooled,
    Mean_diff_ll = Mean_diff - qt(0.975, df)*diff_se,
    Mean_diff_ul = Mean_diff + qt(0.975, df)*diff_se
  ) %>%
  pivot_longer(
    cols=contains("Mean"),
    names_to = "group",
    values_to = "mean"
  )



p1.d <- plot.dat %>%
  filter(group %in% c("Tumor Mean","Non-Tumor Mean")) %>%
  mutate(
    group = ifelse(group == "Tumor Mean", "Tumor", "Non-Tumor"),
    col = ifelse(step == 1, "grey90", "white"),
    h=1, w=Inf
    
  )
p1 <- ggplot()+
  geom_tile(data = p1.d,
              aes(y = description, x=0,
                  height=h, width=w),
              fill = p1.d$col, color=p1.d$col)+
    geom_bar(data=p1.d,
             aes(x=mean, y=description,
                 group=group, color=group,
                 fill=group),
             stat="identity",position = "dodge",
             alpha = 1)+
    labs(x="Mean RelAbundance")+
    theme_classic()+
    theme(
      legend.position = "bottom",
      plot.margin = unit(c(1,0,1,1), "lines")
    )

p2.d <- plot.dat %>%
  filter(group %in% c("Mean_diff", "Mean_diff_ll", "Mean_diff_ul")) %>%
  pivot_wider(
    names_from = group,
    values_from = mean
  ) %>%
  mutate(
    group = ifelse(Mean_diff > 0, "Tumor", "Non-Tumor"),
    p = sprintf("%.3f", round(fdr_p,3)),
    ll = min(Mean_diff_ll)-0.01,
    ul = max(Mean_diff_ul)+0.01
  )
p2<-ggplot(p2.d, aes(x=Mean_diff, y=description))+
    geom_tile(data = p1.d,
              aes(y = description, x=0,
                  height=h, width=w),
              fill = p1.d$col, color=p1.d$col)+
    geom_vline(xintercept = 0, linetype="dashed", alpha=0.5)+
    geom_segment(aes(x=Mean_diff_ll, y=description, xend=Mean_diff_ul, yend=description))+
    geom_point(aes(fill=group, color=group))+
    geom_text(aes(label=p, x=unique(ul)+0.04))+
    coord_cartesian(xlim = c(unique(p2.d$ll), unique(p2.d$ul)),
                    clip = 'off') +
    annotate("text", x=unique(p2.d$ul)+0.08,y = 25,
             angle=90,
             label="q-value (FDR Corrected p-value)")+
    labs(x="Mean Difference in RelAbundance")+
    theme_classic()+
    theme(
      legend.position = "bottom",
      axis.title.y = element_blank(),
      axis.text.y = element_blank(),
      axis.line.y = element_blank(),
      axis.ticks.y = element_blank(),
      plot.margin = unit(c(1,4,1,0), "lines")
    )

# plot
p <- p1 + p2 +
  plot_annotation(title="Stratefied EC Data - Streptococcus sanguinis : 50 most abundant descriptions")
p
ggsave("output/updated-figures-2021-05-20/picrust-strepto-EC.pdf",p,units="in", height=12, width=10)
```


## Modeling Difference between Tumor and Non-Tumor

For the modeling, we used a generalized linear mixed model (GLMM).

First, we looked at the biserial correlation between the abundance of each description and the tumor status.

```{r ec-corplot, warning=F, error=F, message=F, fig.dim=c(7,5), out.height="400px", out.width="100%"}

tb <- mydata %>%
  group_by(description)%>%
  summarise(
    r = cor(tumor, Abundance)
  ) %>%
  mutate(
    M=mean(r)
  )

ggplot(tb, aes(x=r))+
  geom_density()+
  geom_vline(aes(xintercept = M))+
  labs(x="Biserial Correlation",title="Relationship between description abundance and tumor (tumor vs. non-tumor)")+
  theme(panel.grid = element_blank())

```



Next, we need the data to be on an interpretable scale.
First, let's use the raw abundance like scale.

```{r ec-glm-overview, warning=F, error=F, message=F, fig.dim=c(7,5), out.height="400px", out.width="100%"}

p <- ggplot(mydata, aes(x=Abundance))+
  geom_density()
p

mydata <- mydata %>%
  mutate(Abundance.dich=ifelse(Abundance==0, 0, 1))
table(mydata$Abundance.dich)

```


Let's first run models by description so that we can avoid the nesting issue initially. We will come back to this to conduct the final model (it will be more powerful).

```{r ec-glm, message=F, eval=T}

DESCRIPTIONS <- unique(mydata$description)
i<-1
dat0 <- mydata %>% filter(description==DESCRIPTIONS[i])

# quasipossion (approximately negative binom) give an approx. answer.
fit0 <- glm(
  Abundance ~ 1 + tumor,
  data= dat0,
  family=quasipoisson(link = "log")
)
summary(fit0)

```

Run over all descriptors

```{r ec-glm-approx}


results.out <- as.data.frame(matrix(ncol=4, nrow=length(DESCRIPTIONS)))
colnames(results.out) <- c("description", "Est", "SE", "p")
#results.out$description <-DESCRIPTIONS

i <- 1
for(i in 1:length(DESCRIPTIONS)){
#for(i in 1:5){ 
  dat0 <- mydata %>%
    filter(description == DESCRIPTIONS[i])
  fit0 <- glm(
  Abundance ~ 1 + tumor,
  data= dat0,
  family=quasipoisson(link = "log")
)
  fit.sum <- summary(fit0)
  results.out[i, 1] <- DESCRIPTIONS[i]
  results.out[i, 2:4] <- fit.sum$coefficients[2, c(1,2,4)]
}

results.out$fdr_p <- p.adjust(results.out$p, method="fdr")

kable(results.out, format="html", digits=3) %>%
  kable_styling(full_width = T)%>%
  scroll_box(width="100%", height="600px")


# merge with tb.ra for full results table
full.res <- left_join(tb.ra, results.out, by = "description")

write.csv(full.res, "output/picrust_ec_stratefied_strepto_data_results.csv", row.names = F)

```


# Part 2 KO Data

```{r pi-dat-ko}

descriptions <- readr::read_tsv("data/PICRUST/ko_pred_metagenome_unstrat_descrip.tsv")[,1:2] %>%
  distinct()
colnames(descriptions) <- c("func", "description")

pi.dat <- readxl::read_xlsx("data/PICRUST-Stratified/KO_strepto.xlsx")
colnames(pi.dat) <- c("func",  colnames(pi.dat)[-c(1)])

# add descriptions to dataframe
pi.dat <- left_join(pi.dat, descriptions, by="func")
pi.dat <- pi.dat[,c(1,161,3:160)]

# aggregate descriptions to get 1 description per row - unique;y defined.
pi.dat <- pi.dat %>%
  group_by(description) %>%
  summarise(across(`X1.S37.Jun172016`:`X99.D01.S37.Jun232016`,.fns = sum))

# make long format
pi.dat <- pi.dat %>%
  pivot_longer(
    cols=`X1.S37.Jun172016`:`X99.D01.S37.Jun232016`,
    names_to = "ID",
    values_to = "Abundance"
  ) %>%
  mutate(
    ID = substr(ID, 2, 99)
  )

d <- as.data.frame(dat.16s)
mydata <- full_join(pi.dat, d)

mydata <- mydata %>%
  mutate(
    ID.n = as.numeric(as.factor(ID)),
    description.n = as.numeric(as.factor(description))
  ) %>%
  group_by(ID) %>%
  mutate(
    dem = ifelse(sum(Abundance)==0, 1, sum(Abundance)),
    RelAbundance = Abundance/dem*100) %>%
  ungroup()%>%
  group_by(description)%>%
  mutate(avgRA = mean(RelAbundance))


```


## Abundance

```{r ko-ra}


# Run on all descriptions
tb.ra1 <- mydata %>%
  group_by(description) %>%
  summarise(ng = n(),
            Overall.M = mean(RelAbundance),
            Overall.SE = sd(RelAbundance)/sqrt(ng))
tb.ra2m <- mydata %>%
  group_by(description, tumor.cat) %>%
  summarise(M = mean(RelAbundance)) %>%
  pivot_wider(id_cols = description,
              names_from = tumor.cat,
              values_from = M)
tb.ra2se <- mydata %>%
  group_by(description, tumor.cat) %>%
  summarise(ng=n(),SE = sd(RelAbundance)/sqrt(ng)) %>%
  pivot_wider(id_cols = description,
              names_from = tumor.cat,
              values_from = SE)
tb.ra2var <- mydata %>%
  group_by(description, tumor.cat) %>%
  summarise(ng=n(), VAR = var(RelAbundance)) %>%
  pivot_wider(id_cols = description,
              names_from = tumor.cat,
              values_from = VAR)
tb.ra2ng <- mydata %>%
  group_by(description, tumor.cat) %>%
  summarise(ng=n()) %>%
  pivot_wider(id_cols = description,
              names_from = tumor.cat,
              values_from = ng)
tb.ra <- left_join(tb.ra1, tb.ra2m)
tb.ra <- cbind(tb.ra, tb.ra2se[,-1])
tb.ra <- cbind(tb.ra, tb.ra2var[,-1])
tb.ra <- cbind(tb.ra, tb.ra2ng[,-1]) 

colnames(tb.ra) <- c("description", "ng", "Overall Mean", "Overall SE", "Non-Tumor Mean", "Tumor Mean", "Non-Tumor SE", "Tumor SE","Non-Tumor Var", "Tumor Var", "Non-Tumor Ng", "Tumor Ng")
tb.ra <- tb.ra %>%
  arrange(desc(`Overall Mean`))
tb.ra <- tb.ra[, c("description", "Overall Mean", "Overall SE","Tumor Mean","Tumor Var", "Tumor SE","Tumor Ng", "Non-Tumor Mean","Non-Tumor Var", "Non-Tumor SE", "Non-Tumor Ng")]

# compute t-test
tb.ra <- tb.ra %>%
  mutate(
    SEpooled = sqrt(`Tumor Var`/`Tumor Ng` + `Non-Tumor Var`/`Non-Tumor Ng`),
    t = (`Tumor Mean` - `Non-Tumor Mean`)/(SEpooled),
    df = ((`Tumor Var`/`Tumor Ng` + `Non-Tumor Var`/`Non-Tumor Ng`)**2)/(((`Tumor Var`/`Tumor Ng`)**2)/(`Tumor Ng`-1) + ((`Non-Tumor Var`/`Non-Tumor Ng`)**2)/(`Non-Tumor Ng`-1)),
    p = pt(q = abs(t), df=df, lower.tail = F)*2,
    fdr_p = p.adjust(p, method="fdr")
  )



kable(tb.ra, format="html", digits=5, caption="Stratefied KO Data - Streptococcus sanguinis : Average RelAbundance of Each Description (sorted in descending order)") %>%
  kable_styling(full_width = T) %>%
  scroll_box(width = "100%", height="600px")

```


```{r ko-full-plot, warning=F, error=F, message=F, fig.dim=c(12,10), out.height="100%", out.width="90%"}


plot.dat <- tb.ra %>%
  arrange(desc(`Overall Mean`)) %>%
  slice_head(n=50)%>%
  mutate(
    description = fct_reorder(description, `Overall Mean`),
    description = factor(description, levels = levels(description), ordered=T)
  ) %>%
  arrange(description) %>%
  mutate(
    id = 1:n(),
    step = ifelse(id%%2 == 0, 1, 0),
    Mean_diff = `Tumor Mean` - `Non-Tumor Mean`,
    diff_se = SEpooled,
    Mean_diff_ll = Mean_diff - qt(0.975, df)*diff_se,
    Mean_diff_ul = Mean_diff + qt(0.975, df)*diff_se
  ) %>%
  pivot_longer(
    cols=contains("Mean"),
    names_to = "group",
    values_to = "mean"
  )



p1.d <- plot.dat %>%
  filter(group %in% c("Tumor Mean","Non-Tumor Mean")) %>%
  mutate(
    group = ifelse(group == "Tumor Mean", "Tumor", "Non-Tumor"),
    col = ifelse(step == 1, "grey90", "white"),
    h=1, w=Inf
    
  )
p1 <- ggplot()+
  geom_tile(data = p1.d,
              aes(y = description, x=0,
                  height=h, width=w),
              fill = p1.d$col, color=p1.d$col)+
    geom_bar(data=p1.d,
             aes(x=mean, y=description,
                 group=group, color=group,
                 fill=group),
             stat="identity",position = "dodge",
             alpha = 1)+
    labs(x="Mean RelAbundance")+
    theme_classic()+
    theme(
      legend.position = "bottom",
      plot.margin = unit(c(1,0,1,1), "lines")
    )

p2.d <- plot.dat %>%
  filter(group %in% c("Mean_diff", "Mean_diff_ll", "Mean_diff_ul")) %>%
  pivot_wider(
    names_from = group,
    values_from = mean
  ) %>%
  mutate(
    group = ifelse(Mean_diff > 0, "Tumor", "Non-Tumor"),
    p = sprintf("%.3f", round(fdr_p,3)),
    ll = min(Mean_diff_ll)-0.01,
    ul = max(Mean_diff_ul)+0.01
  )
p2<-ggplot(p2.d, aes(x=Mean_diff, y=description))+
    geom_tile(data = p1.d,
              aes(y = description, x=0,
                  height=h, width=w),
              fill = p1.d$col, color=p1.d$col)+
    geom_vline(xintercept = 0, linetype="dashed", alpha=0.5)+
    geom_segment(aes(x=Mean_diff_ll, y=description, xend=Mean_diff_ul, yend=description))+
    geom_point(aes(fill=group, color=group))+
    geom_text(aes(label=p, x=unique(ul)+0.025))+
    coord_cartesian(xlim = c(unique(p2.d$ll), unique(p2.d$ul)),
                    clip = 'off') +
    annotate("text", x=unique(p2.d$ul)+0.05,y = 25,
             angle=90,
             label="q-value (FDR Corrected p-value)")+
    labs(x="Mean Difference in RelAbundance")+
    theme_classic()+
    theme(
      legend.position = "bottom",
      axis.title.y = element_blank(),
      axis.text.y = element_blank(),
      axis.line.y = element_blank(),
      axis.ticks.y = element_blank(),
      plot.margin = unit(c(1,4,1,0), "lines")
    )

# plot
p <- p1 + p2 +
  plot_annotation(title="Stratefied KO Data - Streptococcus sanguinis : 50 most abundant descriptions")
p
ggsave("output/updated-figures-2021-05-20/picrust-strepto-KO.pdf",p,units="in", height=12, width=10)
```


## Modeling Difference between Tumor and Non-Tumor

For the modeling, we used a generalized linear mixed model (GLMM).

First, we looked at the biserial correlation between the abundance of each description and the tumor status.

```{r ko-corplot, warning=F, error=F, message=F, fig.dim=c(7,5), out.height="400px", out.width="100%"}

tb <- mydata %>%
  group_by(description)%>%
  summarise(
    r = cor(tumor, Abundance)
  ) %>%
  mutate(
    M=mean(r)
  )

ggplot(tb, aes(x=r))+
  geom_density()+
  geom_vline(aes(xintercept = M))+
  labs(x="Biserial Correlation",title="Relationship between description abundance and tumor (tumor vs. non-tumor)")+
  theme(panel.grid = element_blank())

```



Next, we need the data to be on an interpretable scale.
First, let's use the raw abundance like scale.

```{r ko-glm-overview, warning=F, error=F, message=F, fig.dim=c(7,5), out.height="400px", out.width="100%"}

p <- ggplot(mydata, aes(x=Abundance))+
  geom_density()
p

mydata <- mydata %>%
  mutate(Abundance.dich=ifelse(Abundance==0, 0, 1))
table(mydata$Abundance.dich)

```


Let's first run models by description so that we can avoid the nesting issue initially. We will come back to this to conduct the final model (it will be more powerful).

```{r ko-glm, message=F, eval=T}

DESCRIPTIONS <- unique(mydata$description)
i<-1
dat0 <- mydata %>% filter(description==DESCRIPTIONS[i])

# quasipossion (approximately negative binom) give an approx. answer.
fit0 <- glm(
  Abundance ~ 1 + tumor,
  data= dat0,
  family=quasipoisson(link = "log")
)
summary(fit0)

```

NExt, let's model the percent relative abundance.
This will allow us to make inference about the difference average relative abundance.
Which is a simpler interpretation than trying to model differences in the log relative abundance which will need to be interpreted as the multiplicative relative change.

```{r ko-glm-approx}


results.out <- as.data.frame(matrix(ncol=4, nrow=length(DESCRIPTIONS)))
colnames(results.out) <- c("description", "Est", "SE", "p")
#results.out$description <-DESCRIPTIONS

i <- 1
for(i in 1:length(DESCRIPTIONS)){
#for(i in 1:5){ 
  dat0 <- mydata %>%
    filter(description == DESCRIPTIONS[i])
  fit0 <- glm(
  Abundance ~ 1 + tumor,
  data= dat0,
  family=quasipoisson(link = "log")
)
  fit.sum <- summary(fit0)
  results.out[i, 1] <- DESCRIPTIONS[i]
  results.out[i, 2:4] <- fit.sum$coefficients[2, c(1,2,4)]
}

results.out$fdr_p <- p.adjust(results.out$p, method="fdr")

kable(results.out, format="html", digits=3) %>%
  kable_styling(full_width = T)%>%
  scroll_box(width="100%", height="600px")


# merge with tb.ra for full results table
full.res <- left_join(tb.ra, results.out, by = "description")

write.csv(full.res, "output/picrust_ko_stratefied_strepto_data_results.csv", row.names = F)


```

