---
title: "Results Output for Question 3"
output:
  workflowr::wflow_html:
    toc: true
editor_options:
  chunk_output_type: console
---

# Question 3

```
Q3: Is fuso associated with tumor stage (pTNM) in either data set? Does X bacteria predict stage? Multivariable w/ age, sex, BMI, history of Barrett's
```

Add to this analysis:

* Fusobacterium nucleatum
* Streptococcus sanguinis
* Campylobacter concisus 
* Prevotella spp. 


TCGA drop "not reported" from tumor stage.


```{r data, include=FALSE, echo=FALSE, warning=F, error=F, message=F}
# load packages
source("code/load_packages.R")

# get data
source("code/get_cleaned_data.R")

theme_set(theme_bw())
# pal = "Sequential"
# scale_colour_discrete <-  function(palname=pal, ...){
#   scale_colour_brewer(palette=palname, ...)
# }
# scale_fill_discrete <-  function(palname=pal, ...){
#   scale_fill_brewer(palette=palname, ...)
# }

knitr::opts_chunk$set(out.width = "225%", out.height = 10)


# transform to relative abundances
# phylo.data.nci.umd <- transform_sample_counts(phylo.data.nci.umd, function(x){x / sum(x)})
# phylo.data.tcga.RNAseq <- transform_sample_counts(phylo.data.tcga.RNAseq, function(x){x / sum(x)})
# phylo.data.tcga.WGS <- transform_sample_counts(phylo.data.tcga.WGS, function(x){x / sum(x)})

# melt data down for use
dat.16s <- psmelt(phylo.data.nci.umd)
dat.rna <- psmelt(phylo.data.tcga.RNAseq)
dat.wgs <- psmelt(phylo.data.tcga.WGS)

# fix otu formatting
dat.rna$otu2 <- "a"
dat.wgs$otu2 <- "a"
i <- 1
for(i in 1:nrow(dat.rna)){
  dat.rna$otu2[i] <- str_split(dat.rna$OTU[i], ";")[[1]][7]
}
for(i in 1:nrow(dat.wgs)){
  dat.wgs$otu2[i] <- str_split(dat.wgs$OTU[i], ";")[[1]][7]
}

# set up variables
dat.16s$sample_type <- 0
dat.16s$sample_type[dat.16s$tissue=="T" &
                      dat.16s$Histology=="ADC" &
                      dat.16s$Barretts. == "Y"] <- "EAC tissues w/ Barretts History"
dat.16s$sample_type[dat.16s$tissue=="N" &
                      dat.16s$Histology=="ADC" &
                      dat.16s$Barretts. == "Y"] <- "EAC-adjacent tissue w/ Barretts History"
dat.16s$sample_type[dat.16s$tissue=="BO" &
                      dat.16s$Histology=="Barrets only"&
                      dat.16s$Barretts. == "Y"] <-"Barretts Only"


dat.rna$sample_type <- 0
dat.rna$sample_type[(dat.rna$morphology=="8140/3" |
                       dat.rna$morphology=="8480/3") &
                       dat.rna$SampleType_Level2=="Tumor" &
                       dat.rna$Barrett.s.Esophagus.Reported=="Yes"] <- "EAC tissues w/ Barretts History"
dat.rna$sample_type[(dat.rna$morphology=="8140/3" | 
                       dat.rna$morphology=="8480/3") &
                       dat.rna$SampleType_Level2=="Normal"&
                       dat.rna$Barrett.s.Esophagus.Reported=="Yes"] <- "EAC-adjacent tissue w/ Barretts History"

dat.rna$EACcomp <- 0
dat.rna$EACcomp[(dat.rna$morphology=="8140/3" |
                       dat.rna$morphology=="8480/3") &
                       dat.rna$SampleType_Level2=="Tumor" &
                       dat.rna$Barrett.s.Esophagus.Reported=="Yes"] <- "EAC tissues w/ Barretts History"
dat.rna$EACcomp[(dat.rna$morphology=="8140/3" |
                       dat.rna$morphology=="8480/3") &
                       dat.rna$SampleType_Level2=="Tumor" &
                       dat.rna$Barrett.s.Esophagus.Reported=="No"] <- "EAC tissues w/o Barretts History"


dat.wgs$sample_type <- 0
dat.wgs$sample_type[(dat.wgs$morphology=="8140/3" |
                       dat.wgs$morphology=="8480/3") &
                       dat.wgs$SampleType_Level2=="Tumor" &
                       dat.wgs$Barrett.s.Esophagus.Reported=="Yes"] <- "EAC tissues w/ Barretts History"
dat.wgs$sample_type[(dat.wgs$morphology=="8140/3" | 
                       dat.wgs$morphology=="8480/3") &
                       dat.wgs$SampleType_Level2=="Normal"&
                       dat.wgs$Barrett.s.Esophagus.Reported=="Yes"] <- "EAC-adjacent tissue w/ Barretts History"

dat.wgs$EACcomp <- 0
dat.wgs$EACcomp[(dat.wgs$morphology=="8140/3" |
                       dat.wgs$morphology=="8480/3") &
                       dat.wgs$SampleType_Level2=="Tumor" &
                       dat.wgs$Barrett.s.Esophagus.Reported=="Yes"] <- "EAC tissues w/ Barretts History"
dat.wgs$EACcomp[(dat.wgs$morphology=="8140/3" |
                       dat.wgs$morphology=="8480/3") &
                       dat.wgs$SampleType_Level2=="Tumor" &
                       dat.wgs$Barrett.s.Esophagus.Reported=="No"] <- "EAC tissues w/o Barretts History"


# make tumor vs normal variable
dat.16s$tumor.cat <- factor(dat.16s$tissue, levels=c("BO", "N", "T"), labels = c("Non-Tumor", "Non-Tumor", "Tumor"))
dat.rna$tumor.cat <- factor(dat.rna$SampleType_Level2, levels=c("Normal", "Tumor"), labels = c("Non-Tumor", "Tumor"))
dat.wgs$tumor.cat <- factor(dat.wgs$SampleType_Level2, levels=c("Normal", "Tumor"), labels = c("Non-Tumor", "Tumor"))

# dataset id
dat.16s$source <- "16s"
dat.rna$source <- "rna"
dat.wgs$source <- "wgs"

# plotting ids
dat.16s$X <- paste0(dat.16s$source, "-", dat.16s$tumor.cat)
dat.rna$X <- paste0(dat.rna$source, "-", dat.rna$tumor.cat)
dat.wgs$X <- paste0(dat.wgs$source, "-", dat.wgs$tumor.cat)

# relabel as (0/1) for analysis
dat.16s$tumor <- as.numeric(factor(dat.16s$tissue, levels=c("BO", "N", "T"), labels = c("Non-Tumor", "Non-Tumor", "Tumor"))) - 1
dat.rna$tumor <- as.numeric(factor(dat.rna$SampleType_Level2, levels=c("Normal", "Tumor"), labels = c("Non-Tumor", "Tumor"))) - 1
dat.wgs$tumor <- as.numeric(factor(dat.wgs$SampleType_Level2, levels=c("Normal", "Tumor"), labels = c("Non-Tumor", "Tumor"))) - 1

# presence- absence
dat.16s$pres <- ifelse(dat.16s$Abundance > 0, 1, 0)
dat.16s$pres[is.na(dat.16s$pres)] <- 0
dat.rna$pres <- ifelse(dat.rna$Abundance > 0, 1, 0)
dat.rna$pres[is.na(dat.rna$pres)] <- 0
dat.wgs$pres <- ifelse(dat.wgs$Abundance > 0, 1, 0)
dat.wgs$pres[is.na(dat.wgs$pres)] <- 0

# tumor stage variable
dat.16s$pTNM[is.na(dat.16s$pTNM)] <- "not reported"
dat.16s$tumor.stage <- factor(
  dat.16s$pTNM,
  levels = c("0", "1", "I", "IIA", "IIB", "III", "IV","not reported"),
  labels = c("0", "1", "I", "II", "II", "III", "IV", "not reported")
)

dat.rna$tumor.stage <- factor(
  dat.rna$tumor_stage,
  levels = c("stage i", "stage ia", "stage ib", "stage iib", "stage iia", "stage iii", "stage iiia", "stage iiib", "stage iiic", "stage iv", "stage iva", "not reported"),
  labels = c("I", "I", "I", "II", "II", "III", "III", "III", "III", "IV", "IV", "not reported")
)

dat.wgs$tumor_stage[is.na(dat.wgs$tumor_stage)] <- "not reported"
dat.wgs$tumor.stage <- factor(
  dat.wgs$tumor_stage,
  levels = c("stage i", "stage ia", "stage ib", "stage iib", "stage iia", "stage iii", "stage iiia", "stage iiib", "stage iiic", "stage iv", "stage iva", "not reported"),
  labels = c("I", "I", "I", "II", "II", "III", "III", "III", "III", "IV", "IV", "not reported")
)


# subset to fuso. nuc. only
# Streptococcus sanguinis 
# Campylobacter concisus
# Prevotella spp.

dat.16s.s <- filter(
  dat.16s,
  OTU %in% c(
    "Fusobacterium_nucleatum",
    unique(dat.16s$OTU[dat.16s$OTU %like% "Streptococcus_"]),
    unique(dat.16s$OTU[dat.16s$OTU %like% "Campylobacter_"]),
    "Prevotella_melaninogenica")
)
dat.rna.s <- filter(
  dat.rna,
  otu2 %in% c(
    "Fusobacterium nucleatum",
    "Streptococcus sanguinis",
      "Streptococcus oralis",
      "Streptococcus mitis", 
      "Streptococcus pneumoniae", 
      "Streptococcus parasanguinis", 
      "Streptococcus salivarius",
    "Campylobacter concisus",
    "Prevotella melaninogenica")
)
dat.wgs.s <- filter(
  dat.wgs,
  otu2 %in% c(
    "Fusobacterium nucleatum",
    "Streptococcus sanguinis",
      "Streptococcus oralis",
      "Streptococcus mitis", 
      "Streptococcus pneumoniae", 
      "Streptococcus parasanguinis", 
      "Streptococcus salivarius",
    "Campylobacter concisus",
    "Prevotella melaninogenica")
)

# new names
dat.16s.s$OTU1 <- factor(
  dat.16s.s$OTU,
  levels = c(
    "Fusobacterium_nucleatum",
    "Streptococcus_dentisani:Streptococcus_infantis:Streptococcus_mitis:Streptococcus_oligofermentans:Streptococcus_oralis:Streptococcus_pneumoniae:Streptococcus_pseudopneumoniae:Streptococcus_sanguinis",
    "Campylobacter_rectus:Campylobacter_showae",
    "Prevotella_melaninogenica"
  ),
  labels = c(
    "Fusobacterium nucleatum",
    "Streptococcus sanguinis",
    "Campylobacter concisus",
    "Prevotella melaninogenica"
  )
)
dat.rna.s$OTU1 <- factor(
  dat.rna.s$otu2,
  levels = c(
    "Fusobacterium nucleatum",
    "Streptococcus sanguinis",
    "Campylobacter concisus",
    "Prevotella melaninogenica"),
  labels = c(
    "Fusobacterium nucleatum",
    "Streptococcus sanguinis",
    "Campylobacter concisus",
    "Prevotella melaninogenica")
)

dat.wgs.s$OTU1 <- factor(
  dat.wgs.s$otu2,
  levels = c(
    "Fusobacterium nucleatum",
    "Streptococcus sanguinis",
    "Campylobacter concisus",
    "Prevotella melaninogenica"),
  labels = c(
    "Fusobacterium nucleatum",
    "Streptococcus sanguinis",
    "Campylobacter concisus",
    "Prevotella melaninogenica")
)

# rename bacteria
dat.16s.s$OTU <- factor(
  dat.16s.s$OTU,
  levels = c(
    "Fusobacterium_nucleatum",
    "Streptococcus_dentisani:Streptococcus_infantis:Streptococcus_mitis:Streptococcus_oligofermentans:Streptococcus_oralis:Streptococcus_pneumoniae:Streptococcus_pseudopneumoniae:Streptococcus_sanguinis",
    "Campylobacter_rectus:Campylobacter_showae",
    "Prevotella_melaninogenica"
  ),
  labels = c(
    "Fusobacterium nucleatum",
    "Streptococcus spp.",
    "Campylobacter concisus",
    "Prevotella melaninogenica"
  )
)
dat.rna.s$OTU <- factor(
  dat.rna.s$otu2,
  levels = c(
    "Fusobacterium nucleatum",
    "Streptococcus sanguinis",
      "Streptococcus oralis",
      "Streptococcus mitis", 
      "Streptococcus pneumoniae", 
      "Streptococcus parasanguinis", 
      "Streptococcus salivarius",
    "Campylobacter concisus",
    "Prevotella melaninogenica"),
  labels = c(
    "Fusobacterium nucleatum",
    "Streptococcus spp.",
      "Streptococcus spp.",
      "Streptococcus spp.", 
      "Streptococcus spp.", 
      "Streptococcus spp.", 
      "Streptococcus spp.",
    "Campylobacter concisus",
    "Prevotella melaninogenica")
)

dat.wgs.s$OTU <- factor(
  dat.wgs.s$otu2,
  levels = c(
    "Fusobacterium nucleatum",
    "Streptococcus sanguinis",
      "Streptococcus oralis",
      "Streptococcus mitis", 
      "Streptococcus pneumoniae", 
      "Streptococcus parasanguinis", 
      "Streptococcus salivarius",
    "Campylobacter concisus",
    "Prevotella melaninogenica"),
  labels = c(
    "Fusobacterium nucleatum",
    "Streptococcus spp.",
      "Streptococcus spp.",
      "Streptococcus spp.", 
      "Streptococcus spp.", 
      "Streptococcus spp.", 
      "Streptococcus spp.",
    "Campylobacter concisus",
    "Prevotella melaninogenica")
)

dat.rna$OTU <- dat.rna$otu2
dat.wgs$OTU <- dat.wgs$otu2

```



# NCI 16s data

## Double Checking Data

```{r nci-fig1, fig.height=5, out.width="100%", out.height="100%"}

# in long format
table(dat.16s$tumor.stage)
# by subject
dat <- dat.16s %>% filter(OTU == "Fusobacterium_nucleatum")
table(dat$tumor.stage)
sum(table(dat$tumor.stage)) # sample size met


dat.16s.s$Abundancej <- dat.16s.s$Abundance+0.01
mean.dat <- dat.16s.s %>%
  group_by(tumor.stage, OTU) %>%
  summarize(M = mean(Abundance))
ggplot(dat.16s.s, aes(x=tumor.stage, y=Abundancej))+
  geom_violin()+
  geom_jitter(alpha=0.25,width = 0.25)+
  geom_point(data=mean.dat, aes(x=tumor.stage, y = M), size=2, alpha =0.9, color="red")+
  labs(x="Tumor Stage", 
       title="Distribution of abundance across tumor stage",
       subtitle="Red dot is average abundnace")+
  scale_y_continuous(
    breaks=c(0, 10, 100, 200, 300, 400, 500),
    limits = c(0,500),
    trans="sqrt")+
  facet_wrap(.~OTU, nrow=1)+
  theme_classic()

```


Stage "1" has only 1 unique sample and will be dropped from subsequent analyses.

```{r}
dat.16s.s <- dat.16s.s %>%
  filter(tumor.stage != "1")%>%
  mutate(tumor.stage = droplevels(tumor.stage, exclude="1"))
```

## Multinomial Logistic Regression

### Model 1: TS ~ Abundance

```{r glm-16s-m1}

fit <- nnet::multinom(tumor.stage ~ OTU, data=dat.16s.s)
summary(fit)

# obtain significance tests
z <- summary(fit)$coefficients/summary(fit)$standard.errors
z
p <- (1 - pnorm(abs(z), 0, 1)) * 2
p
## extract the coefficients from the model and exponentiate
exp(coef(fit)) # interpret as relative risks
# save fitted logits
pp <- fitted(fit)
# preditive data
dotu <- data.frame(OTU = c("Fusobacterium nucleatum", "Streptococcus spp.", "Campylobacter concisus", "Prevotella melaninogenica"))
predict(fit, newdata = dotu, "probs") # only TINY differences

## store the predicted probabilities for each value of ses
pp.otu <-cbind(dotu, predict(fit, newdata = dotu, "probs", se = TRUE))

## calculate the mean probabilities within each level of OTU
by(pp.otu[, 2:7], pp.otu$OTU, colMeans)
```



### Model 2: TS ~ OTU + Abundance

```{r glm-16s-m2, fig.height=5, out.width="100%", out.height="100%"}

fit <- nnet::multinom(tumor.stage ~ OTU + Abundance, data=dat.16s.s)
summary(fit)

# obtain significance tests
z <- summary(fit)$coefficients/summary(fit)$standard.errors
z
p <- (1 - pnorm(abs(z), 0, 1)) * 2
p
## extract the coefficients from the model and exponentiate
exp(coef(fit)) # interpret as relative risks
# save fitted logits
pp <- fitted(fit)
# predit data
dotu <- data.frame(OTU = c("Fusobacterium nucleatum", "Streptococcus spp.", "Campylobacter concisus", "Prevotella melaninogenica"), Abundance = mean(dat.16s.s$Abundance))
predict(fit, newdata = dotu, "probs") # bigger differences

## look at the averaged predicted probabilities for different values of the continuous predictor variable Abundnace within each level of OTU
dabund <- data.frame(
  OTU = rep(c("Fusobacterium nucleatum", "Streptococcus spp.", "Campylobacter concisus", "Prevotella melaninogenica"), each = 51),
  Abundance = rep(seq(0, 500,10), 4)
)
pp.abund <-cbind(dabund, predict(fit, newdata = dabund, "probs", se = TRUE))

## calculate the mean probabilities within each level of OTU
by(pp.abund[, 3:8], pp.abund$OTU, colMeans)

## melt data set to long for ggplot2
lpp <- melt(pp.abund, id.vars = c("OTU", "Abundance"), value.name = "probability")

## plot predicted probabilities across Abundance values for each level of OTU
## facetted by tumor.stage
ggplot(lpp, aes(x = Abundance, y = probability, colour = OTU)) +
  geom_line() + 
  facet_grid(variable ~., scales="free")+
  labs(y="Probability of Tumor Stage",
       title="Tumor stage likelihood with bacteria abundance and OTU")+
  theme(
    panel.grid = element_blank()
  )
```


### Model 3: TS ~ OTU + Abundance + OTU:Abundnace

```{r glm-16s-m3, fig.height=5, out.width="100%", out.height="100%"}

fit <- nnet::multinom(tumor.stage ~ OTU + Abundance + OTU:Abundance, data=dat.16s.s)
summary(fit)

# obtain significance tests
z <- summary(fit)$coefficients/summary(fit)$standard.errors
z
p <- (1 - pnorm(abs(z), 0, 1)) * 2
p
## extract the coefficients from the model and exponentiate
exp(coef(fit)) # interpret as relative risks
# save fitted logits
pp <- fitted(fit)
# predit data
gmeans <- dat.16s.s %>% group_by(OTU) %>% summarise(M = mean(Abundance))
dotu <- data.frame(OTU = c("Fusobacterium nucleatum", "Streptococcus spp.", "Campylobacter concisus", "Prevotella melaninogenica"), Abundance = gmeans$M)
predict(fit, newdata = dotu, "probs") # bigger differences

## look at the averaged predicted probabilities for different values of the continuous predictor variable Abundnace within each level of OTU
dabund <- data.frame(
  OTU = rep(c("Fusobacterium nucleatum", "Streptococcus spp.", "Campylobacter concisus", "Prevotella melaninogenica"), each = 51),
  Abundance = rep(seq(0, 500,10), 4)
)
pp.abund <-cbind(dabund, predict(fit, newdata = dabund, "probs", se = TRUE))

## calculate the mean probabilities within each level of OTU
by(pp.abund[, 3:8], pp.abund$OTU, colMeans)

## melt data set to long for ggplot2
lpp <- melt(pp.abund, id.vars = c("OTU", "Abundance"), value.name = "probability")

## plot predicted probabilities across Abundance values for each level of OTU
## facetted by tumor.stage
ggplot(lpp, aes(x = Abundance, y = probability, colour = OTU)) +
  geom_line() + 
  facet_grid(variable ~., scales="free")+
  labs(y="Probability of Tumor Stage",
       title="Tumor stage likelihood with bacteria abundance and OTU")+
  theme(
    panel.grid = element_blank()
  )
```



# TCGA RNAseq data

## Double Checking Data

```{r rna-fig1, fig.height=5, out.width="100%", out.height="100%"}

# in long format
table(dat.rna$tumor.stage)
# by subject
dat <- dat.rna %>% filter(OTU == "Fusobacterium nucleatum")
table(dat$tumor.stage)
sum(table(dat$tumor.stage)) # sample size met

dat.rna.s$Abundancej <- dat.rna.s$Abundance+0.01
mean.dat <- dat.rna.s %>%
  group_by(tumor.stage, OTU) %>%
  summarize(M = mean(Abundance))
ggplot(dat.rna.s, aes(x=tumor.stage, y=Abundancej))+
  geom_violin()+
  geom_jitter(alpha=0.25,width = 0.25)+
  geom_point(data=mean.dat, aes(x=tumor.stage, y = M), size=2, alpha =0.9, color="red")+
  labs(x="Tumor Stage", 
       title="Distribution of abundance across tumor stage",
       subtitle="Red dot is average abundnace")+
  scale_y_continuous(
    breaks=c(0, 100, 1000, 10000, 100000, 150000),
    #limits = c(0,500),
    trans="pseudo_log")+
  facet_wrap(.~OTU, nrow=1)+
  theme_classic()
```

## Multinomial Logistic Regression

### Model 1: TS ~ Abundance

```{r glm-rna-m1}

fit <- nnet::multinom(tumor.stage ~ OTU, data=dat.rna.s)
summary(fit)

# obtain significance tests
z <- summary(fit)$coefficients/summary(fit)$standard.errors
z
p <- (1 - pnorm(abs(z), 0, 1)) * 2
p
## extract the coefficients from the model and exponentiate
exp(coef(fit)) # interpret as relative risks
# save fitted logits
pp <- fitted(fit)
# preditive data
dotu <- data.frame(OTU = c("Fusobacterium nucleatum", "Streptococcus spp.", "Campylobacter concisus", "Prevotella melaninogenica"))
predict(fit, newdata = dotu, "probs") # only TINY differences

## store the predicted probabilities for each value of ses
pp.otu <-cbind(dotu, predict(fit, newdata = dotu, "probs", se = TRUE))

## calculate the mean probabilities within each level of OTU
by(pp.otu[, 2:6], pp.otu$OTU, colMeans)
```



### Model 2: TS ~ OTU + Abundance

```{r glm-rna-m2, fig.height=5, out.width="100%", out.height="100%"}

fit <- nnet::multinom(tumor.stage ~ OTU + Abundance, data=dat.rna.s)
summary(fit)

# obtain significance tests
z <- summary(fit)$coefficients/summary(fit)$standard.errors
z
p <- (1 - pnorm(abs(z), 0, 1)) * 2
p
## extract the coefficients from the model and exponentiate
exp(coef(fit)) # interpret as relative risks
# save fitted logits
pp <- fitted(fit)
# predit data
dotu <- data.frame(OTU = c("Fusobacterium nucleatum", "Streptococcus spp.", "Campylobacter concisus", "Prevotella melaninogenica"), Abundance = mean(dat.rna.s$Abundance))
predict(fit, newdata = dotu, "probs") # bigger differences

## look at the averaged predicted probabilities for different values of the continuous predictor variable Abundnace within each level of OTU
dabund <- data.frame(
  OTU = rep(c("Fusobacterium nucleatum", "Streptococcus spp.", "Campylobacter concisus", "Prevotella melaninogenica"), each = 51),
  Abundance = rep(seq(0, 500,10), 4)
)
pp.abund <-cbind(dabund, predict(fit, newdata = dabund, "probs", se = TRUE))

## calculate the mean probabilities within each level of OTU
by(pp.abund[, 3:7], pp.abund$OTU, colMeans)

## melt data set to long for ggplot2
lpp <- melt(pp.abund, id.vars = c("OTU", "Abundance"), value.name = "probability")

## plot predicted probabilities across Abundance values for each level of OTU
## facetted by tumor.stage
ggplot(lpp, aes(x = Abundance, y = probability, colour = OTU)) +
  geom_line() + 
  facet_grid(variable ~., scales="free")+
  labs(y="Probability of Tumor Stage",
       title="Tumor stage likelihood with bacteria abundance and OTU")+
  theme(
    panel.grid = element_blank()
  )
```


### Model 3: TS ~ OTU + Abundance + OTU:Abundnace

```{r glm-rna-m3, fig.height=5, out.width="100%", out.height="100%"}

fit <- nnet::multinom(tumor.stage ~ OTU + Abundance + OTU:Abundance, data=dat.rna.s)
summary(fit)

# obtain significance tests
z <- summary(fit)$coefficients/summary(fit)$standard.errors
z
p <- (1 - pnorm(abs(z), 0, 1)) * 2
p
## extract the coefficients from the model and exponentiate
exp(coef(fit)) # interpret as relative risks
# save fitted logits
pp <- fitted(fit)
# predit data
gmeans <- dat.rna.s %>% group_by(OTU) %>% summarise(M = mean(Abundance))
dotu <- data.frame(OTU = c("Fusobacterium nucleatum", "Streptococcus spp.", "Campylobacter concisus", "Prevotella melaninogenica"), Abundance = gmeans$M)
predict(fit, newdata = dotu, "probs") # bigger differences

## look at the averaged predicted probabilities for different values of the continuous predictor variable Abundnace within each level of OTU
dabund <- data.frame(
  OTU = rep(c("Fusobacterium nucleatum", "Streptococcus spp.", "Campylobacter concisus", "Prevotella melaninogenica"), each = 51),
  Abundance = rep(seq(0, 500,10), 4)
)
pp.abund <-cbind(dabund, predict(fit, newdata = dabund, "probs", se = TRUE))

## calculate the mean probabilities within each level of OTU
by(pp.abund[, 3:7], pp.abund$OTU, colMeans)

## melt data set to long for ggplot2
lpp <- melt(pp.abund, id.vars = c("OTU", "Abundance"), value.name = "probability")

## plot predicted probabilities across Abundance values for each level of OTU
## facetted by tumor.stage
ggplot(lpp, aes(x = Abundance, y = probability, colour = OTU)) +
  geom_line() + 
  facet_grid(variable ~., scales="free")+
  labs(y="Probability of Tumor Stage",
       title="Tumor stage likelihood with bacteria abundance and OTU")+
  theme(
    panel.grid = element_blank()
  )
```



# TCGA WGS data

## Double Checking Data

```{r wgs-fig1, fig.height=5, out.width="100%", out.height="100%"}

# in long format
table(dat.wgs$tumor.stage)
# by subject
dat <- dat.wgs %>% filter(OTU == "Fusobacterium nucleatum")
table(dat$tumor.stage)
sum(table(dat$tumor.stage)) # sample size met

dat.wgs.s$Abundancej <- dat.wgs.s$Abundance+0.01
mean.dat <- dat.wgs.s %>%
  group_by(tumor.stage, OTU) %>%
  summarize(M = mean(Abundance))
ggplot(dat.wgs.s, aes(x=tumor.stage, y=Abundancej))+
  geom_violin()+
  geom_jitter(alpha=0.25,width = 0.25)+
  geom_point(data=mean.dat, aes(x=tumor.stage, y = M), size=2, alpha =0.9, color="red")+
  labs(x="Tumor Stage", 
       title="Distribution of abundance across tumor stage",
       subtitle="Red dot is average abundnace")+
  scale_y_continuous(
    breaks=c(0, 100, 1000, 10000, 100000),
    #limits = c(0,500),
    trans="pseudo_log")+
  facet_wrap(.~OTU, nrow=1)+
  theme_classic()
```

## Multinomial Logistic Regression

### Model 1: TS ~ Abundance

```{r glm-wgs-m1}

fit <- nnet::multinom(tumor.stage ~ OTU, data=dat.wgs.s)
summary(fit)

# obtain significance tests
z <- summary(fit)$coefficients/summary(fit)$standard.errors
z
p <- (1 - pnorm(abs(z), 0, 1)) * 2
p
## extract the coefficients from the model and exponentiate
exp(coef(fit)) # interpret as relative risks
# save fitted logits
pp <- fitted(fit)
# preditive data
dotu <- data.frame(OTU = c("Fusobacterium nucleatum", "Streptococcus spp.", "Campylobacter concisus", "Prevotella melaninogenica"))
predict(fit, newdata = dotu, "probs") # only TINY differences

## store the predicted probabilities for each value of ses
pp.otu <-cbind(dotu, predict(fit, newdata = dotu, "probs", se = TRUE))

## calculate the mean probabilities within each level of OTU
by(pp.otu[, 2:6], pp.otu$OTU, colMeans)
```



### Model 2: TS ~ OTU + Abundance

```{r glm-wgs-m2, fig.height=5, out.width="100%", out.height="100%"}

fit <- nnet::multinom(tumor.stage ~ OTU + Abundance, data=dat.wgs.s)
summary(fit)

# obtain significance tests
z <- summary(fit)$coefficients/summary(fit)$standard.errors
z
p <- (1 - pnorm(abs(z), 0, 1)) * 2
p
## extract the coefficients from the model and exponentiate
exp(coef(fit)) # interpret as relative risks
# save fitted logits
pp <- fitted(fit)
# predit data
dotu <- data.frame(OTU = c("Fusobacterium nucleatum", "Streptococcus spp.", "Campylobacter concisus", "Prevotella melaninogenica"), Abundance = mean(dat.wgs.s$Abundance))
predict(fit, newdata = dotu, "probs") # bigger differences

## look at the averaged predicted probabilities for different values of the continuous predictor variable Abundnace within each level of OTU
dabund <- data.frame(
  OTU = rep(c("Fusobacterium nucleatum", "Streptococcus spp.", "Campylobacter concisus", "Prevotella melaninogenica"), each = 51),
  Abundance = rep(seq(0, 500,10), 4)
)
pp.abund <-cbind(dabund, predict(fit, newdata = dabund, "probs", se = TRUE))

## calculate the mean probabilities within each level of OTU
by(pp.abund[, 3:7], pp.abund$OTU, colMeans)

## melt data set to long for ggplot2
lpp <- melt(pp.abund, id.vars = c("OTU", "Abundance"), value.name = "probability")

## plot predicted probabilities across Abundance values for each level of OTU
## facetted by tumor.stage
ggplot(lpp, aes(x = Abundance, y = probability, colour = OTU)) +
  geom_line() + 
  facet_grid(variable ~., scales="free")+
  labs(y="Probability of Tumor Stage",
       title="Tumor stage likelihood with bacteria abundance and OTU")+
  theme(
    panel.grid = element_blank()
  )
```


### Model 3: TS ~ OTU + Abundance + OTU:Abundnace

```{r glm-wgs-m3, fig.height=5, out.width="100%", out.height="100%"}

fit <- nnet::multinom(tumor.stage ~ OTU + Abundance + OTU:Abundance, data=dat.wgs.s)
summary(fit)

# obtain significance tests
z <- summary(fit)$coefficients/summary(fit)$standard.errors
z
p <- (1 - pnorm(abs(z), 0, 1)) * 2
p
## extract the coefficients from the model and exponentiate
exp(coef(fit)) # interpret as relative risks
# save fitted logits
pp <- fitted(fit)
# predit data
gmeans <- dat.wgs.s %>% group_by(OTU) %>% summarise(M = mean(Abundance))
dotu <- data.frame(OTU = c("Fusobacterium nucleatum", "Streptococcus spp.", "Campylobacter concisus", "Prevotella melaninogenica"), Abundance = gmeans$M)
predict(fit, newdata = dotu, "probs") # bigger differences

## look at the averaged predicted probabilities for different values of the continuous predictor variable Abundnace within each level of OTU
dabund <- data.frame(
  OTU = rep(c("Fusobacterium nucleatum", "Streptococcus spp.", "Campylobacter concisus", "Prevotella melaninogenica"), each = 51),
  Abundance = rep(seq(0, 500,10), 4)
)
pp.abund <-cbind(dabund, predict(fit, newdata = dabund, "probs", se = TRUE))

## calculate the mean probabilities within each level of OTU
by(pp.abund[, 3:7], pp.abund$OTU, colMeans)

## melt data set to long for ggplot2
lpp <- melt(pp.abund, id.vars = c("OTU", "Abundance"), value.name = "probability")

## plot predicted probabilities across Abundance values for each level of OTU
## facetted by tumor.stage
ggplot(lpp, aes(x = Abundance, y = probability, colour = OTU)) +
  geom_line() + 
  facet_grid(variable ~., scales="free")+
  labs(y="Probability of Tumor Stage",
       title="Tumor stage likelihood with bacteria abundance and OTU")+
  theme(
    panel.grid = element_blank()
  )
```




